Negative cell cycle regulation by Calcineurin is necessary for proper beta cell regeneration in zebrafish

Version 1

Abstract

Abstract

Stimulation of pancreatic beta cell regeneration could be a therapeutic lead to treat diabetes. Unlike humans, the zebrafish can efficiently regenerate beta cells, notably from ductal pancreatic progenitors. To gain insight into the molecular pathways involved in this process, we established the transcriptomic profile of the ductal cells after beta cell ablation in the adult zebrafish. These data highlighted the protein phosphatase calcineurin as a new potential modulator of beta cell regeneration. We showed that calcineurin overexpression abolished the regenerative response, leading to glycemia dysregulation. On the opposite, calcineurin inhibition increased ductal cell proliferation and subsequent beta cell regeneration. Interestingly, the enhanced proliferation of the progenitors was paradoxically coupled with their exhaustion. This suggests that the proliferating progenitors are next entering in differentiation. Calcineurin appears as a guardian which prevents an excessive progenitor proliferation to preserve the pool of progenitors. Altogether, our findings reveal calcineurin as a key player in the balance between proliferation and differentiation to enable a proper beta cell regeneration.

Introduction

Blood glucose homeostasis is tightly controlled by pancreatic endocrine cells. Insulin-producing beta cells work in close association with alpha cells, which secrete glucagon, and delta cells secreting somatostatin, to maintain normal glycemia. The insulin plays a critical role in this process as it is the only hormone able to lower the glycemia. Beta cell loss is a hallmark of type 1 and of late stages of type 2 diabetes, leading to chronic hyperglycemia. Beta cell destruction is largely irreversible in human and in mammals, making the disease incurable nowadays. Nevertheless, studies in diabetic mice models uncovered promising evidences of slight recovery of beta cells via regeneration. For example, new beta cells can arise from the replication of remaining beta cells [1]. Different models of pancreas injuries revealed the plasticity of mammalian pancreatic cells. Actually, differentiated endocrine cells such as alpha cells[2] and delta cells[3] can reprogram and convert into insulin-producing cells, a process that is age dependent. Another possible source of beta cells could be de novo formation from pancreatic progenitors residing in the ductal compartment of the adult pancreas [4]. Lineage tracing experiments could not clearly highlight new beta cells arising from the ductal cells in the adult [4]–[6]. However, expression of the pro-endocrine marker Neurog3 was detected in the ducts in different mouse models of pancreas regeneration [4], [7], [8]. In addition, a rare population of ductal cells expressing Neurog3 has been reported to contribute to beta cell neogenesis during diabetes [9]. A recent study of single cell RNA sequencing reveal a sub-population of human ductal cells which are able to give rise to all the pancreatic cell types, including beta cells, following implantation in mice [10]. Altogether, these findings suggest that mammalian pancreatic ducts possess the intrinsic capacity to (re)generate beta cells even though this process is poorly efficient and slow, especially in adults. In contrast to mammals, zebrafish possess remarkable capacity of regeneration, independently of its age [11]–[16]. This model can therefore be exploited to identify and characterize regenerative mechanisms and ultimately induce regeneration in mammals. Based on regenerative mechanisms identified in zebrafish, several studies succeed to improve regeneration in mammals, underlying the possibility of translation from zebrafish to mammals ([17]–[21] and [22], for review).

In zebrafish, the nitroreductase (NTR)/nitroaromatic prodrug system is widely used. In this technique, the combination of cell type-specific NTR expression and nitroaromatic prodrug exposure allow for controlled and targeted cell ablation [23], [24]. When NTR is expressed under the control of the insulin (ins) promoter (Tg(ins:NTR-mCherry) [25]), beta cell destruction is complete within 3 days following the treatment in adult fish, which correlate with a peak of hyperglycemia [26], [27]. The regeneration of beta cells upon ablation is spontaneous and fast and the glycemia is normalized within two weeks [13], [29]. Similar to mice, new beta cells can arise through proliferation of surviving beta cells [28] as well as through the contribution of alpha cells [29], [30] and delta cells[31], [32], underscoring the overall conservation of these processes from zebrafish to mammals. Nevertheless, unlike mammals, the presence of pancreatic progenitors in the ducts is well established in both larval [33] and adult zebrafish [26], [27]. Lineage tracing experiments pointed out that ductal cells and centroacinar cells (CACs) give rise to new beta cells [26], [33]. In adults, duct-derived beta cells start to be detected between 7 and 10 days following beta cell ablation [26], [27].

Notch signaling is a key player regulating the differentiation of duct-associated progenitors into endocrine cells. Larval duct cells as well as adult CACs display strong Notch activity [26], [34]. This signaling pathway has a central role in beta cell genesis during both development [35] and regeneration [33] by repressing endocrine differentiation. In zebrafish, different levels of Notch activity determine the behavior of the pancreatic progenitors. While a high level of Notch activity maintains cells in quiescence, a moderate level induces the entry in the cell cycle and proliferation whereas a low level drives endocrine differentiation of the progenitor cells [35]. A steep decrease of Notch activity pushes the progenitors to differentiate prematurely, bypassing the amplification step and leading to their depletion [35]. Repression of the Notch signaling by mTor, activated by glucose and nutrients, hence promotes beta cell formation and regeneration from ductal progenitors [33]. While Notch and mTor signaling are crucial for this process, there is still a need to establish a global view of the molecular mechanisms regulating beta cell regeneration.

To identify early events regulating ductal-derived beta cell regeneration, we determined the transcriptomic signature of ductal cells from adult zebrafish following beta cell destruction. Our data highlighted an upregulation of the Calcineurin (CaN) pathway. To elucidate CaN function in beta cell regeneration, we both repressed and activated CaN pathway. We showed that CaN regulates beta cell neogenesis in the ducts during regeneration, by modulating progenitor proliferation. Together, our findings underline that CaN fine tunes the balance between progenitor proliferation and beta cell differentiation to guarantee proper regeneration.

Results

Transcriptomic profiling of ductal cells after beta cell destruction highlights regulation of calcineurin pathway

To gain a better understanding of the molecular mechanisms underlying the regeneration of beta cells from the ducts, we determined the transcriptional landscape of ductal cells by RNA-sequencing after beta cell ablation in adults. To selectively ablate the beta cells, we used the Tg(ins:NTR-mCherry) transgenic fish and the ductal cells were labelled through the Tg(nkx6.1:GFP) reporter line [27] in which GFP marks the ductal tree and associated multipotent pancreatic progenitors [27]. More precisely, 3 to 4 fishes were treated with the pro-drug MTZ at 10mM overnight, to induce ablation of beta cells or with DMSO for the non-ablated controls. Ablation was confirmed by blood glucose measurement before collection of the pancreas. To capture the early events triggered by the destruction of beta cells, we generated the transcriptome of the ducts 3 days post ablation treatment (dpt), i.e. before beta cell neogenesis. Differential gene expression analysis revealed that 1866 genes are up-regulated and 1515 genes down-regulated in the ductal cell of fish treated with MTZ compared to control (pAdj < 0.05). According to Gene Ontology analysis, the most enriched pathways among the up-regulated genes were DNA replication and cell cycle (Figure 1A). This further corroborates our previous findings and those of others regarding the activation of duct-associated progenitors’ proliferation in response to beta cell ablation[26], [27]. As expected, the Notch pathway was enriched in the down-regulated genes [33], [35] (Figure 1B). Intriguing in a context of strong proliferative response, we found an enrichment of the cellular senescence signature (Figure 1A). We looked further in the genes associated with this signature. Among them were found several components of the calcineurin signaling pathway such as nfatc3b, ppp3ccb (the catalytic subunit of Calcineurin), itpr2 and calm3b (Figure 1C). In addition to these genes related to cellular senescence, our RNAseq studies revealed the modulation of other genes from the CaN canonical pathway (Figure 1D, genes in green), underlying its potential role in beta cell regeneration. Calcineurin (CaN) is a highly conserved calcium/calmodulin dependent Ser/Thr phosphatase, involved in numerous biological process including fin regeneration and beta cell function, [36]–[39]. This prompted us to investigate the role of CaN in beta cell regeneration Previous RNAseq data performed in our laboratory indicate that CaN (ppp3cca/b) and NFATc3 (nfatc3a/b) are mainly express in endocrine cells [40] (Figure 1A supplemental), which is in accordance with the role of CaN/NFAT signaling in beta cells [41]. CaN genes (ppp3cca/b) as well as nfatc3b, are express at lower levels in the ducts at basal state but their expression is induced in response to beta cell destruction (Figure 1B supplemental).

Transcriptomic profiling of ductal cells during beta cell regeneration and validation in larvae

A-B) Enrichment ratio of selected non-redundant signatures of KEGG pathways overrepresented in ductal cells after beta cells ablation (UP – A and DOWN-B) compared to ductal cells without beta cells ablation. Gene Ontology (GO) terms were identified using ORA analysis by WebGestalt (Liao et al., 2019) using the list of DE genes provided by DESeq. The light color for Notch pathway means p-value = 0.11.

C) List of genes associated with the signature of cellular senescence from A-B. Genes related to CaN pathway are in bold.

D) Calcineurin canonical pathway with genes in green that are modulated in transcriptomic data from A-B.

E) Experimental design for regeneration test in larvae. Briefly, after nifurpirinol treatment from 3 to 4dpf, larvae were fixed and analyzed at 4-7-10 and 14 days post treatment (dpt).

F) Graph representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) at 0-4-7-10 and 14 dpt. The gray spheres represent non-ablated conditions and the pink triangles the ablated condition. Data are presented as mean values ± SEM. One-way ANOVA test with Tukey’s multiple comparison test, ****p-value>0.0005.

G) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) larvae at 14dpt. 3D projection (stack) of one non-ablated and one ablated representative samples. The principal islet (PI) and the pancreatic tail are shown. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 100µM.

H) Experimental design for EdU assay in larvae. After NFP treatment for 3 to 4dpf, larvae were exposed to EdU at 2dpt before fixation for analysis.

I) Whole mount fluorescent immunohistochemistry (GFP and EdU) of the pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) larvae at 2dpt. 3D projection (stack) of one non-ablated and one ablated representative samples. Arrows point out GFP+ duct cells EdU+ in the pancreatic tail. Scale 50µM.

J) Barplot representing the percentage of GFP+ ductal cells which incorporated EdU+ in non-ablated and ablated conditions. Data are presented as mean values ± SD. T-test. **p-value>0.005.

Calcineurin activity regulates the ductal regenerative response

To evaluate the role of CaN in beta cell regeneration, more specifically derived from ductal progenitors, we have chosen to use young larvae, where regenerated beta cells in the pancreatic tail arise exclusively from the ducts [33]. In response to beta cells ablation, the intrapancreatic ducts undergo a ductal regenerative response whereby differentiation toward the endocrine fate is increased [33]. We first determined the rate of beta cell neogenesis from the ducts in response to a single acute ablation of beta cells as we performed in adults. We treated Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae with nifurpirinol (NFP) from 3 to 4 dpf mCherry+ beta cells were quantified in the GFP+ ducts present in the tail at several time points: 4, 7, 10 and 14 dpt (Figure 1E). Duct-associated beta cells started to be detected in non-ablated larvae between 7 and 10 dpt (Figure 1F) and the number of beta cells slowly increased until 14 dpt (Figure 1F, 1G). In ablated larvae, the increase became more pronounced from 10 dpt onwards (Figure 1F, 1G), indicating faster endocrine differentiation. This experiment establishes that the ductal regenerative response is detectable between 10 and 14 days after the ablation of beta cells, performed at 3 dpf. We next wanted to determine if ductal cell proliferation is activated in response to beta cell destruction in larvae as in adult fish. We exposed Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae to EdU the second day following ablation (Figure 1H). In ablated larvae, the proportion of GFP+ ductal cells EdU+ (in S-phase) was higher compared to non-ablated larvae (Figure 1I, 1J). This result shows that acute beta cell ablation in larvae rapidly activates ductal cell proliferation, as previously reported in adult zebrafish.

As our transcriptomic data from adult zebrafish revealed modulation of CaN pathway at 3 dpt, we treated Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae from 1 to 3 dpt after beta cell ablation, with a CaN inhibitor, the Cyclosporin A (CsA) [39] (Figure 2A). The number of newly formed beta cells was monitored in the tail from 4 to 14 dpt. CsA enhanced beta cell formation at 10 dpt (Figure 2B-D). However, this effect appears to be transient since no discernible difference was observed between the control and CsA-treated larvae in regeneration at the latest time point, suggesting an acceleration of the regenerative response (Figure 2B, Figure 2A supplemental). Interestingly, CsA did not affect beta cell differentiation in non-ablated larvae indicating that CsA only acts in a regenerative context (Figure 2B-D). Of note, CsA increased as well the number of regenerated beta cells in the principal islet (Figure 2B supplemental).

Calcineurin inhibition with CsA increases the ductal regenerative response

A) Experimental design for regeneration test in larvae with CsA treatment. Briefly, after nifurpirinol treatment from 3 to 4dpf, larvae were treated with CsA from 1 to 3dpt and fixed and analyzed at 4-7-10 and 14 days post treatment (dpt).

B) Graph representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) at 0-4-7-10 and 14 dpt. The gray spheres represent non-ablated condition ; the pink triangles represent the ablated condition ; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SEM. Two-way ANOVA test with Sidak’s multiple comparisons test, *p-value>0.05.

C) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) larvae at 10dpt. 3D projection (stack) of non-ablated and ablated larvae treated with DMSO or CsA representative samples. The principal islet (PI) and the pancreatic tail are shown. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 100µM.

D) Barplot representing the number of number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) larvae at 10dpt. The gray spheres represent non-ablated condition ; the pink triangles represent the ablated condition ; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SD. Two-way ANOVA with Tukey’s multiple comparison test, *p-value>0.05.

E) Graph representing the mean number of GFP+ neurod1+ cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(neurod1:GFP) at 0-4-7 and 10 dpt. The gray spheres represent non-ablated condition ; the pink triangles represent the ablated condition ; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SEM. Two-way ANOVA test with Sidak’s multiple comparisons test, **p-value>0.005; *** p-value>0.0005; **** p-value>0.00005.

F) Graph representing the mean number of GFP+ neurod1 EdU+ cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(neurod1:GFP) at 0-4-7 and 10 dpt. The gray spheres represent non-ablated condition ; the pink triangles represent the ablated condition ; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SEM.

We next tested the effect of CsA on endocrine progenitors in a regenerative context. We induced regeneration in Tg(neurod1:GFP) larvae where the GFP is expressed in both endocrine progenitors and mature endocrine cells. We first assessed generation of GFP cells at different time points (Figure 2D supplemental) and showed that CsA induced an increase of neurod1+ cells from 4 dpt. The increase was still detectable at least until 10dpt (Figure 2E, Figure 2E supplemental). To determine if these additional cells result from their own proliferation, we performed a pulse of EdU just before analysis (Figure 2D supplemental). We observed that CsA did not affect the neurod1+ proliferation rate that is very low at these stages (Figure 2F). As a consequence, the effect of CsA cannot be explained by endocrine cell proliferation but rather by neogenesis from progenitors. As CsA affects pro-endocrine cells formation, we next wondered if the increased cell formation induced by CsA is specific to beta cells. Treatment with CsA was performed as previously and delta and alpha cells were detected by immunofluorescence. Interestingly, CsA did not affect alpha nor delta1.1 cells neogenesis in response to beta cell ablation (>Figure 2F-G supplemental). Overall, these experiments showed that CsA affects specifically the beta cells and their endocrine progenitors.

CaN over-activation abolishes the regenerative response

We then wondered if an opposite regulation of CaN i.e. its activation impacts as well the regenerative response. To that end, we generated a transgenic line Tg(hsp70:GFP-P2A-ppp3ccCA) that allows ubiquitous expression of a constitutively active form of CaNCA (ppp3ccCA) upon heat shocks. Beta cell ablation was triggered in Tg(hsp70:GFP-P2A-ppp3ccCA); Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae from 3 to 4 dpf and CaNCA expression was induced by four successive heat-shocks from 1 to 3 dpt (Figure 3A). The overexpression of CaNCA after ablation impaired the regenerative response at 14 dpt (Figure 3B-C). Similar results were obtained with Tg(UAS:GFP-P2A-ppp3ccCA); Tg(cftr:gal4)[45] ; Tg(ins:NTR-P2A-mCherry) larvae in which CaNCA is constitutively and specifically overexpressed in the ducts within the pancreas (Figure 3D-E). Importantly, the structure of the ducts was similar in CaNCA overexpressing larvae compared to Tg(nkx6.1:GFP) controls (Figure 3D), suggesting that the suppression of the regenerative response in CaNCA-overexpressing larvae was not due to morphogenetic defects during ductal growth. This result strongly suggest that CaN acts directly in the ducts to regulate beta cell regeneration while it is not necessary for normal beta cell differentiation.

Transgenic mediated overexpression of calcineurin abolish the ductal regenerative response

A) Experimental design for regeneration test in larvae with heat-shocks. Briefly, after nifurpirinol treatment from 3 to 4dpf, four heat shock were performed from 1 to 3dpt and larvae were fixed and analyzed at 14dpt.

B) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(hsp70:CaNCA); Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 14dpt. 3D projection (stack) of one non-ablated and one ablated with or without heat-shock representative samples. The principal islet (PI) and the pancreatic tail are showed. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 100µM.

C) Barplot representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(hsp70:CaNCA); Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 14dpt. The gray spheres represent non-ablated condition ; the pink triangles represent the ablated condition ; the black squares heat-shock condition and inverted green triangles ablated + heat-shock condition. Data are presented as mean values ± SD. Two-way ANOVA with Tukey’s multiple comparisons test, *p-value>0.05, *** p-value>0.0005, ns = non-significant.

D) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of larvae at 14dpt. 3D projection (stack) of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) one non-ablated and one ablated representative control samples and Tg(UAS:CaNCA); Tg(ins:NTR-P2A-mCherry) ; Tg(cftr:gal4) one non-ablated and one ablated representative samples. The principal islet (PI) and the pancreatic tail are showed. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 100µM.

E) Barplot representing the mean number of mCherry+ beta cells in the pancreatic tail of larvae at 14dpt. The gray spheres represent non-ablated Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) condition ; the pink triangles represent the ablated Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) condition ; the black squares non-ablated Tg(UAS:CaNCA); Tg(ins:NTR-P2A-mCherry) ; Tg(cftr:gal4) condition and inverted green triangles ablated Tg(UAS:CaNCA); Tg(ins:NTR-P2A-mCherry) ; Tg(cftr:gal4) condition. Data are presented as mean values ± SD. Two-way ANOVA with Tukey’s multiple comparison test, *** means p-value>0.0005, **** p-value>0.00005, ns = non-significant.

CaN regulates beta cell differentiation induced by Notch inhibition in absence of regeneration

Our transcriptomic data showed that the Notch pathway is downregulated in ductal cells during beta cell regeneration (Figure 1). As the level of Notch activity determines the behavior of ductal cells [33] from quiescence to proliferation and subsequently to beta cell differentiation, we tested whether calcineurin acts together with the Notch pathway on a common pool of ductal progenitors. To inhibit the Notch pathway, we treated larvae with several concentrations of the gamma-secretase inhibitor LY411575 from 3 to 4dpf in absence of regeneration. The activity of CaN was inhibited by CsA during the same timeframe (Figure 4A). As previously, we used reporter lines for beta and ductal cells Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) and the number of secondary beta cells was analyzed at 6dpf (Figure 4A). As expected, the number of beta cells progressively rose as the concentration of the Notch inhibitor increased (Figure 4B). Combined treatment with CsA fostered the differentiation of beta cells between 1 and 10 µM LY411575 but did not result in further increase at 15 µM LY411575 (Figure 4B), suggesting that CaN is important within a permissive window of Notch activity. Since we overserved the highest synergistic effect at 5µM of LY411575, we used this concentration for the following experiments (Figure 4C-D). It is worth noting that combined treatment of LY411575 (5uM) and another CaN inhibitor, FK506, resulted in the same synergistic increase of beta cell differentiation (Figure 4A-B supplemental), confirming that the effect is due to CaN inhibition. The combined effect of Notch and CaN inhibition is transient as it is not observed at 7dpf anymore (Figure 4E). Therefore, as observed in regenerative conditions (Figure 2B), CaN inhibition accelerates beta cell neogenesis induced by Notch repression.

CaN repression potentializes the effect of Notch inhibition on beta cell formation

A) Experimental design for Notch inhibition test in non-ablated condition. Larvae were treated concomitantly with LY411575 (Notch inhibitor) and CsA from 3 to 4dpf and were fixed and analyzed at 6dpf.

B) Graph representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) larvae at 6 dpf depending the concentration of LY411575. The blue dots represent LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SEM. Two-way ANOVA test with Sidak’s multiple comparison test, *p-value>0.05, **p-value>0.05, ns = non-significant.

C) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) larvae at 6dpf. 3D projection (stack) of one control (without any treatment) ; one CsA-treated ; one LY411757-treated and one with both CsA and LY411575 treated larvae. The principal islet (PI) and the pancreatic tail are showed. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 50µM.

D) Barplot representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) larvae at 6dpf. The black dots represent the control; gray CsA treatment ;blue LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SD. Two-way ANOVA with Tukey multiple comparison test, ***p-value>0.0005, ****p-value>0.00005.

E) Graph representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) larvae at 5-6-7 dpf. The black dots represent the control; gray CsA treatment ;blue LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SEM. Two-way ANOVA test with Sidak’s multiple comparison test, *p-value>0.05.

To determine to which extent CaN pathway can modulate Notch induced beta cell neogenesis, we activated CaNCA overexpression in Tg(hsp70:GFP-P2A-ppp3ccCA); Tg(ins:NTR-P2A-mCherry) larvae by an heat-shock at 3 dpf and treated them with LY411575 (Figure 4C supplemental). CaNCA overexpression resulted in a lowered beta cell formation induced by Notch inhibition (Figure 4D supplemental), revealing that CaN activation counterbalanced the effects of Notch inhibition. Using our previous settings of Notch and CaN inhibition, we next wondered if the canonical pathway downstream of CaN was involved in the enhancement of beta cell differentiation. To activate NFAT, a well-known target of CaN, we used CHIR99021 allowing a stabilization of the active form of NFAT (Figure 1D). We found that CHIR99021 rescued the effect of CsA (Figure 4D supplemental), suggesting that CaN inhibition increases beta cell neogenesis at least partially by the regulation of NFAT. Overall, these results reveal that CaN regulates beta cell formation in pro-endocrinogenic context, such as induced by a low level of Notch activity. Moreover, it suggests that both CaN and Notch pathway act on a common pool of ductal progenitors to regulate beta cell neogenesis.

CaN controls the proliferation of duct-associated progenitors induced by Notch inhibition

Given that we observed the most significant increase in beta cell formation with CsA when Notch activity was mildly repressed (Figure 4B-D), and since mild Notch activity has been shown to promote progenitor amplification [35], it suggests that CaN acts at this level. To explore this possibility, we exposed briefly Tg(nkx6.1:GFP) larvae to EdU after mild Notch (LY411575 5µM) and CaN inhibition (Figure 5A). As expected, Notch inhibition increased the proportion of proliferating EdU+ GFP+ duct cells at 4dpf (Figure 5B-C) and decreased the amount of ductal progenitors 2 days later (Figure 5D), which is concomitant with the increase of beta cell differentiation (Figure 4C-D). Interestingly, the combined inhibition with CsA further increased these proportions (Figure 5B-D). In contrast, CaN inhibition alone had no effect on basal proliferation or on the number of ductal progenitors (Figure 5B-D). Incidentally, at 4dpf, while the proliferation is increased (Figure 5C), the number of ductal cells remained the same in all conditions (Figure 5A supplemental), suggesting that ductal cells have not yet left the cell cycle to differentiate. In comparison, after stronger Notch inhibition (15µM), the ductal cells are already depleted at 4dpf (Figure 5B supplemental), as they directly differentiate without entering the cell cycle [35]. In these conditions, CsA could therefore not enhance ductal progenitor proliferation and thus beta cell formation (Figure 4B). These results show that CaN and Notch pathways act together on the proliferation of the ductal progenitors to prevent their exhaustion.

CaN repression increases the proportion of duct proliferating cells

A) Experimental design for EdU assay in Notch test. Larvae were treated concomitantly with LY411575 (Notch inhibitor) and CsA from 3 to 4dpf and then briefly treated with EdU before fixation and analysis at 4 dpf or at 6dpf.

B) Whole mount fluorescent immunohistochemistry (GFP and EdU) of the pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) larvae at 4dpf. 3D projection (stack) of one control (without any treatment), one with CsA only, one with LY411575 only and one with both CsA and LY411757 representative samples. Scale 50µM.

C) Barplot representing the percentage of GFP+ ductal cells which incorporated EdU+ in pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) larvae for the Notch test. The black dots represent the control; gray CsA treatment ;blue LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SD. T-test. Two-way ANOVA test with Tukey’s multiple comparisons test, *p-value>0.05 ; ****p-value>0.00005 ; ns = non-significant.

D) Barplot representing the number of GFP+ ductal cells which in pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) larvae at 6dpf for the Notch test. The black dots represent the control; gray CsA treatment ;blue LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SD. T-test. Two-way ANOVA test with Tukey’s multiple comparisons test, *p-value>0.05 ; ****p-value>0.00005 ; ns = non-significant.

E) Experimental design for EdU assay in regeneration. Larvae were treated with nifurpirinol for beta cell ablation from 3 to 4dpf then with CsA from 4 to 5dpf and then briefly treated with EdU before fixation and analysis.

F) Whole mount fluorescent immunohistochemistry (GFP and EdU) of the pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) larvae at 5dpf. 3D projection (stack) of one representative sample of non-ablated or ablated with or without CsA are shown. Scale 50µM.

G) Barplot representing the percentage of GFP+ ductal cells which incorporated EdU+ in pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) larvae at 5dpf. The gray spheres represent non-ablated condition ; the pink triangles the ablated condition ; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SD. Two-way ANOVA test with Tukey’s multiple comparisons test, *p-value>0.05 ; ****p-value>0.00005 ; ns = non-significant.

H) Whole mount fluorescent immunohistochemistry (VenusPest and EdU) of the pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(tp1:VenusPest) larvae at 5dpf. 3D projection (stack) of one representative sample of non-ablated or ablated with or without CsA are shown. Scale 50µM.

I) Barplot representing the percentage of GFP+ ductal cells which incorporated EdU+ in pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(tp1:VenusPest) larvae at 5dpf. The gray spheres represent non-ablated condition ; the pink triangles the ablated condition ; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± sd. Two-way ANOVA test with Tukey multiple comparisons test, *p-value>0.05 ; ****p-value>0.00005 ; ns means non-significant.

J) Barplot representing the number of VenusPest+ ductal cells which incorporated EdU+ in pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(tp1:VenusPest) larvae at 5dpf. The gray spheres represent non-ablated condition ; the pink triangles the ablated condition ; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SD. Two-way ANOVA test with Tukey’s multiple comparisons test, **p-value>0.005.

CaN prevents exhaustion of Notch responsive progenitors during beta cell regeneration

Taken together, our results indicate that CaN plays a role in regulating the proliferation of ductal progenitors in contexts that are permissive for beta cell differentiation. To demonstrate that CaN regulates ductal progenitor proliferation in a similar manner to Notch inhibition but during regeneration, we exposed Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae after ablation and CsA treatment (Figure 5E). CaN inhibition enhanced the proliferation of ductal cells in ablated larvae (Figure 5F-G). To next determine if CaN acts on the Notch-responsive progenitors in beta cell regeneration, we then used Tg(tp1:VenusPest) Notch reporter line. At larval stages the vast majority of ductal cells are Notch responsive and the reporter line marks all the progenitors within the ductal tree [34]. We treated Tg(ins:NTR-P2A-mCherry); Tg(tp1:VenusPest) as described above (Figure 5E). In ablated-larvae, CsA increased tp1+ ductal cell proliferation (Figure 5H-I). Moreover, CsA induced a reduction of tp1+ ductal cells in ablated larvae (Figure 5 H-J), suggesting an exhaustion of the Notch responsive progenitors, in accordance with premature beta cell differentiation we observed (Figure 2B-4D). Lastly, CsA does not affect tp1+ cells in non-ablated larvae showing that CaN inhibition did not directly affect Notch signaling (Figure 5J). Those results suggest that CaN fine tunes the balance between proliferation of the progenitors and their differentiation to prevent their exhaustion during beta cell regeneration.

CaN regulation is functionally relevant in adult zebrafish

To further expand upon our findings and investigate their relevance in a non-developmental context, we next ought to determine whether CaN function is maintained in older zebrafish. We used 2 months old Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) juveniles fish to perform beta cell ablation followed by CsA treatments. We analyzed the number of small islets (up to 5 cells) at 7 and 10dpt. At 7dpt, CsA increased the number of small islets in ablated larvae (Figure 6A-B) showing that CaN inhibition enhances beta cell regeneration in juvenile zebrafish, as in larvae. It is noteworthy that, as in larvae, we highlighted an acceleration of beta cell regeneration. The increase number of small islets is indeed transient as it not observe anymore at 10dpt (Figure 6C).

CaN regulation is important in juveniles/adults and necessary for correct glycemia recovery

A) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) 2 months old zebrafish at 7dpt. 3D projection (stack) of non-ablated and ablated larvae treated with DMSO or CsA representative samples. The principal islet (PI) and the pancreatic tail are shown. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 200µM

B-C) Barplot representing the number of number of mCherry+ small secondary islets (=<5 cells) in the pancreatic tail of Tg(ins:NTR-P2A-mCherry) ; Tg(nkx6.1:GFP) 2 months old zebrafish at 7 (B) and 10dpt (C). The gray spheres represent non-ablated condition ; the pink triangles represent the ablated condition ; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SD. Two-way ANOVA with Tukey’s multiple comparison test, *p-value>0.05.

D-E) Barplot representing the glycemia (mg/dL) of Tg(ins:NTR-P2A-mCherry) ; months old adult zebrafish at 7 (D) and 10dpt (E). The pink triangles represent the ablated condition ; the inverted green triangles ablated + CsA condition; the blue squares Tg(hsp70:CaNCA) after heat shocks ; the orange lozenges Tg(UAS:CaNCA); Tg(cftr:gal4). The grey line represents the mean glycemia of controls (non-ablated) fish. Data are presented as mean values ± SD. One-way ANOVA with Tukey’s multiple comparison test, *p-value>0.05; **p-value>0.005.

Next, we determined the functional impact of CaN overexpression or inhibition by assessing the glycemia at 7, 10 and 14 after beta cell ablation. Overexpression of CaNCA using either Tg(hsp70:GFP-P2A-ppp3ccCA) or Tg(cftr:gal4); Tg(UAS:GFP-P2A-ppp3ccCA), led to an increased glycemia at both 7 and 10dpt (Figure 6D-E), before recovery at 14 dpt (Figure 6A supplemental). This indicates that the overexpression of CaN impedes the recovery of glycemia induced by beta cell regeneration. However, CaN inhibition did not seem to further improve the glycemia (Figure 6D-E) probably because the glycemia was already low at 7dpt (77 mg/dL on average) compared to non-ablated control (mg/dl) (Figure 6D-E). Altogether, these results show that in adult zebrafish also, CaN regulation is necessary to enable beta cell regeneration and for proper recovery of the glycemia after beta cell loss.

Discussion

The present study reveals the importance of the protein phosphatase CaN as a new player in beta cell regeneration in zebrafish. We demonstrate that CaN act on ductal-associated progenitor cells by balancing proliferation and endocrine differentiation. In addition, we show the interplay between CaN and Notch signalling, a master regulator of beta cell regeneration.

Previous drug and genetic screening using zebrafish larvae enabled the identification of several regulators of beta cell regeneration from different pancreatic cellular sources. For example, adenosine has been shown to stimulate beta cell replication [28] and, igfbp1a [42] and TGFb suppression [30] promote alpha-to-beta cell transdifferentiation. As for cdk5 inhibition and folinic acid/Folr1, they promote beta cell regeneration from the pancreatic ducts [43], [44]. Here, to identify novel regulators of beta cell regeneration specifically from pancreatic ducts, we carried out a transcriptomic profiling of duct cells following beta cell ablation in the adult zebrafish. Transcriptomic analyses show that the regulated genes encompass most of the genes and pathways identified in previous studies (igfbp1, mTor, Notch, etc.), underlying the importance of those actors in beta cell regeneration. Our data reveal also that DNA replication is the most enriched signature attesting that duct cells undergo a potent proliferative response after the destruction of beta cells.

Besides these expected signatures, our transcriptomic data uncover the unanticipated upregulation of numerous genes implicated in DNA repair and cell cycle arrest. These signatures might indicate that highly proliferating ductal cells activate counteracting mechanisms. Among the genes regulated in those signatures, we focused on CaN and determined its role in beta cell regeneration. Pharmacological inhibition of CaN increases the proliferation of duct cells induced by beta cell ablation, resulting in an acceleration of beta cell regeneration in the ducts. Consistently, transgene-mediated CaN overactivation abolishes the regenerative response. Importantly, the inhibition of regeneration is observed when CaN is overexpressed either ubiquitously or selectively in cftr-expressing ductal cells indicating that the role of CaN in beta cell regeneration is intrinsic to the ducts. Based on functional assays in larvae, we not only confirm the activation of the proliferation of ductal cells soon after beta cell ablation, but also that the rate of progenitor proliferation is carefully controlled by CaN in order to achieve proper and timely regeneration of beta cells. Our data are consistent with earlier studies reporting a role of CaN in proliferation dynamics during fin regeneration. In the regenerating fin, low CaN activity is found in the proximal region of the blastema characterized by a high rate of proliferation and regeneration and its activity increases distally where lower proliferation is observed [36], [38], [39]. It was suggested that CaN control blastemal cell progeny divisions [36]. In human, the importance of CaN in proliferation is also highlighted in organ transplanted patients. When patients are treated with Cyclosporin A (i.e. the CaN inhibitor we used in this study) as immunosuppressive drug they indeed present an increased risk of skin cancer, notably due to keratinocyte senescence inhibition [45].

It has been shown that Notch inhibitory treatments switch progenitors from proliferative self-renewing to premature differentiation, leading to progenitor depletion [34], [35]. Our study reveals that this phenomenon is further exacerbated by CaN inhibition. Importantly, during normal larval development in absence of Notch inhibitory treatment, CaN does not affect basal ductal proliferation nor beta cell differentiation. Hence, Notch signalling has to be repressed to detect the effect of CaN on the progenitors, suggesting that CaN acts downstream of Notch pathway. In differentiating keratinocytes, CaN cooperates with Notch signalling to regulate p21/cdkn1a (which is upregulated in the ducts at 3 dpt), cell cycle withdrawal and differentiation [46]. These studies show that CaN acts in association with Notch signalling on progenitors proliferation and on their differentiation.

Based on our data, we build the model depicted in Figure 7A. Our study suggests that CaN acts in competent progenitor cells and that this competence is determined by Notch signalling. When Notch is repressed to a mild level, the progenitor enter into the cell cycle and acquire a pro-endocrinogenic competence [35]. CaN acts on these progenitors to tone down an excessive proliferation and avoid the exhaustion of these progenitors. CaN is therefore a guardian of the progenitor population. CaN inhibition both increases progenitor proliferation and induces their depletion, suggesting a switch to a symmetric division resulting in two daughter cells entering in endocrine differentiation.

Model of CaN action on ductal progenitors to regenerate beta cells

A) Under physiological conditions, the behavior of the ductal progenitors is determined by Notch signaling. Calcineurin is active in these progenitors and enable a proper control between proliferation and differentiation. When CaN in repressed, more ductal progenitors enter in the cell cycle (2dpt) and switch to a mode of proliferation leading to differentiation of the two daughter cells (4dpt), as more pro-endocrine cells are formed. The result is a exhaustion of the progenitors and a premature beta cell differentiation (10dpt).

More precisely, a previous study in mice uncover the existence of three division modes of the pancreatic progenitors during embryonic development, i.e. symmetric self-renewing resulting in two progenitors cells ; asymmetric resulting in a progenitor and a endocrine cell and symmetric differentiative resulting in two differentiated cells [47]. The authors actually show that the type of division is defined by the timing of induction of endocrine program by NEUROG3 [47]. Interestingly, NEUROG3 seems to be the link between proliferation and differentiation, and its expression is regulated by Notch signalling [48]. In zebrafish, endocrine differentiation is not induced by Neurog3 but by Ascl1b and Neurod1 [49]. Concerning that subject, CaN inhibition accelerates the formation of neurod1+ cells. Hence CaN could possibly act via the determination of the type of division i.e. symmetric vs asymmetric. This model is supported by previous observation in others systems. In stem cells and neuronal and hematopoietic progenitors, premature differentiation results from a switch in the mode of cellular division, from symmetric amplifying division to asymmetric differentiating division [50], [51]. Notch determines the choice between both types of divisions [52], [53].

Calcineurin is known to be implicated in cellular senescence [45]. Usually thought as negative regulators of development and cellular growth, DNA repair [54] and cellular senescence [55] appear to be required in both developmental and regenerative processes. Our transcriptomic data suggest that these mechanisms are required for beta cell regeneration. Therefore, it would be interesting to determine the contribution of these cellular processes in the ductal progenitors and determine if CaN acts via cellular senescence in this case.

Overall, this study brings new insights on beta cell regeneration and highlights the ductal progenitor cell cycle as a cornerstone in the process. Some studies report an increase of proliferation of some ductal cell population in diabetic patients [10], [56], implying a regenerative response. However, these ductal cells cannot efficiently reform the beta cell mass, suggesting a dormant mechanism of regeneration. As such, the balance between proliferation and induction of endocrine differentiation could be a key to improve beta cell neogenesis. However, as CaN is also important for beta cell function, this approach would require to be transient to induce neogenesis. Therefore, it should be combined with methods to induce beta cell proliferation to ultimately reconstitute the beta cell mass. Overall, this study brings a better understanding on the regulation of the balance between ductal progenitors proliferation and endocrine differentiation. These results should provide new hints to help improve regenerative competences in mammals.

Material and Methods

Key resources table

Zebrafish husbandry and generation of the Tg(hsp70:eGFP-P2A-ppp3ccaCA)ulg068 and Tg(UAS:eGFP-P2A-ppp2ccaCA)ulg069 zebrafish lines

Tg BAC(nkx6.1:eGFP)ulg004 [57] ; Tg(ins:NTR-P2A-mCherry)ulg034 [58] ; Tg(cftr:gal4) and Tg(tp1:VenusPest) were used. Zebrafish were raised in standard conditions at 28°C. All experiments were carried out in compliance with the European Union and Belgian law and with the approval of the ULiège Ethical Committee for experiments with laboratory animals (approval number : 2075).

The hsp70:GFP-P2A-ppp3ccaCA transgene has been generated by cloning a PCR fragment containing the Gateway vector pCR8/GW/TOPO. Firstly, we amplified the full length of ppp3ccaCA with primers IM369/IM380 and amplified GFP-P2A with overlapping regions with IM371/O226. The overlapping PCR used the primers O226/IM380. Then to obtain a truncated ppp3ccaCA lacking the calmodulin biding and the autoinhibitory domain, resulting in a constantly active form of calcineurin, we amplified the last fragments with IM370/O226 and cloned into PCR8 vector. The promoter was assembled by LR recombination with pE5-hsp70 into pDestTol2p2A from the Tol2kit [59]. Tg (hsp70:GFP-P2A-ppp3ccaCA) fish have been generated using the Tol2 mediated transgenesis [60]. The Tg(UAS:GFP-P2A-ppp3ccacA) has been generated by ligation (KLD kit, NEB) of PCR fragments GFP-P2A-ppp3ccacA (IM459/IM460) with the UAS sequences (IM457/IM458) in plasmid from (Distel 2009) and then inserted into pDestTol2p2A from the Tol2kit. Final constructions has been injected with transposase into wild type (WT) AB embryos.

Beta-cell ablation

Adults fish for RNA-sequencing experiment were treated with freshly prepared metronidazole (MTZ) (Sigma M1547) at 10mM with 0.2% DMSO in fish water. Control treatments consisted of fish water containing 0.2% DMSO. Fish were treated for 18 hours in the dark. Nifurpirinol (NFP) (32439, Sigma-Aldrich) stock solution was dissolved at 2.5 mM in DMSO. b-cell ablation in Tg(nkx6.1:eGFP); Tg(ins:NTR-P2A-mCherry) larvae was induced by treatment with 2.5 µM NFP in E3. Control treatments consisted of E3 containing 0.2% DMSO. Larvae were treated for 18 hours in the dark.

Drug treatments

Cyclosporine A (Selleckchem, S2286), CHIR99021 and LY411575 (Sigma-Aldrich, SML0506) stock solution were dissolved at 10mM in DMSO. Larvae treatment were respectively performed at 0.1 µM and 5µM in E3. Control treatments consisted of E3 containing the same amount of DMSO than drug treatment. Larvae were treated for 18 hours in the dark.

5-ethynyl-2′-deoxyuridine (EdU) incorporation assay

Zebrafish larvae were incubated in 4 mM EdU dissolved in E3 water for 8 hours, they were then directly euthanised in tricaine and fixed in 4% PFA. EdU was detected according to the protocol of Click-iT™ EdU Cell Proliferation Kit for Imaging, Alexa Fluor™ 647 (ThermoFisher C10340) after whole mount immunodetection.

Heat Shock

Successive heat shocks of 30 minutes and 12 hours apart were performed at 39°C for larvae and 37°C for juveniles and adults zebrafish.

Whole mount immunodetection

Larvae were euthanized in tricaine and fixed in 4% PFA at 4 °C for IHC. The digestive tract of juveniles was dissected prior immunodetection and kept in methanol for at least 18 hours. After depigmentation with 3% H2O2/1% KOH, larvae were permeabilised 30 min in PBS/ Triton X-100 and incubated for two hours in blocking buffer (4% goat serum/1% BSA/PBS/0.1% Triton X-100). Primary and secondary antibodies were incubated at 4 °C overnight.

Primary antibodies: Living Colors Polyclonal anti-mCherry/dsRed (rabbit, 1:500, Clontech 632496), anti-GFP (chicken, 1:1000), Secondary antibodies: Alexa Fluor-488,-568,-633 (goat, 1:750, Molecular Probes).

Flow cytometry and FACS

The whole pancreas from 3-4 fish of Tg(nkx6.1:eGFP); Tg(ins:NTR-P2A-mCherry) adult fish (6–10 months old, males and females) were dissected, collected and washed in HBSS without Ca2+/Mg2+. Live cell dissociation was performed in Tryple Select 1x solution (GIBCO) supplemented with 100 U/mL collagenase IV (Life Technologies 17104-019) and 40 µg/mL proteinase K (Invitrogen, 25530031) for 10 min at 28 °C.

The GFP+ cells were selected on FACS Aria III and sorted under purity mode and after exclusion of the doublets. The purity of the sorted cells was confirmed by epifluorescence microscopy (∼95 %). Cells (about 1000-5000/fish depending on the cell type) were immediately lysed with 0.5% Triton X-100 containing 2U/µl RNAse inhibitor and stored at –80 °C.

mRNA sequencing of FACSed cells and bioinformatic analyses

cDNAs were prepared from lysed cells according to SMART-Seq2.0 (Picelli et al., 2014) for low input RNA sequencing and libraries were prepared with Nextera® DNA Library kit (Illumina). Independent biological replicates of each cell type sequenced using Illumina NextSeq500 and obtained ∼20 million 100 bp paired-end reads. Reads were mapped and aligned to the zebrafish genome GRCz11 from Ensembl gene annotation version using STAR version 2.6.1 (Dobin et al., 2013). Gene expression levels were calculated with featureCounts (http://bioinf.wehi.edu.au/featureCounts/) and differential expression determined with DESeq2 (Love, Huber, & Anders, 2014). Expression values are given as normalized read counts. Poorly expressed genes with mean normalized expression counts <10 were excluded from the subsequent analyses. DESeq2 uses Wald test for significance with posterior adjustment of P values (Padj) using Benjamini and Hochberg multiple testing. The differentially expressed (DE) genes identified with a Padj cutoff of 0.05 were submitted for GO analysis using WebGestalt tool (Liao, Wang, Jaehnig, Shi, &Zhang, 2019).

Confocal microscopy and image analysis

Images were acquired using Leica SP5 confocal microscope. We used ImageJ to count the cells and Imaris to do the pictures.

Glycemia measurement

Glycemia measurement were performed as described in [58].

Data availability

The sequences that support the findings are of this study have been deposit in Geo with the accessions codes GSE212124, https://www.ncbi.nlm.nih.gov/geo/info/linking.html. The authors declare that all other data supporting the findings of this study are available within the paper and its supplementary information files.

Authors contribution

L.M. conducted experiments, acquired and analysed data, analysed transcriptomic data, wrote the manuscript. D.B. conducted RNA-seq experiment and revised the manuscript A.L. implemented bioinformatic programs and analysed transcriptomic data. C.R., C.G. and C.D. conducted experiments. L.F. revised the manuscript. B.P. provided resources and advised on research studies. M.L.V. provided resources, advised on research studies and revised the manuscript I.M. supervised the whole work, acquired funding, designed the research studies, verified data and assisted in writing the manuscript.

eLife Assessment

This work presents some valuable information regarding the molecular mechanisms controlling the regeneration of pancreatic beta cells following induced cell ablation. However, the study lacks the critical lineage tracing result to support the conclusion about the origin of the regenerated beta cells. The results of the pharmacological manipulation of CaN signaling are also incomplete. In particular, these manipulation are not cell-specific, making it difficult to interpret and thus genetic approach is recommended.

Reviewer #1 (Public Review):

Induction of beta cell regeneration is a promising approach for the treatment of diabetes. In this study, Massoz et.al., identified calcineurin (CaN) as a new potential modulator of beta cell regeneration by using zebrafish as model. They also showed that calcineurin (CaN) works together with Notch signaling calcineurin (CaN) to promote the beta cell regeneration. Overall, the paper is well organized, and technically sound. However, some evidence seems weak to get the conclusion.

Reviewer #2 (Public Review):

This work started with transcriptomic profiling of ductal cells to identify the upregulation of calcineurin in the zebrafish after beta-cell ablation. By suppressing calcineurin with its chemical inhibitor cyclosporin A and expressing a constitutively active form of calcineurin ubiquitously or specifically in ductal cells, the authors found that inhibited calcineurin activity promoted beta-cell regeneration transiently while ectopic calcineurin activity hindered beta-cell regeneration in the pancreatic tail. They also showed similar effects in the basal state but only when it was within a particular permissive window of Notch activity. To further investigate the roles of calcineurin in the ductal cells, the authors demonstrated that calcineurin inhibition additionally induced the proliferation of the ductal cells in the regenerative context or under a limited level of Notch activity. Interestingly, the enhanced proliferation was followed by a depletion of ductal cells, suggesting that calcineurin inhibition would exhaust the ductal cells. Based on the data, the authors proposed a very attractive and intriguing model of the role of calcineurin in maintaining the balance of the progenitor proliferation and the endocrine differentiation. However, the conclusions of this paper are only partially supported by the data as some evidence from the data remains suggestive.

1. In the transcriptomic profiling, genes differentially regulated in the ablated adults could be solely due to the chemical effects of metronidazole instead of the beta-cell ablation. A control group without ins:NTR-mCherry but treated with metronidazole is necessary to exclude the side effects of metronidazole.

2. Although it has been shown that the pancreatic duct is a major source of the secondary islets in the pancreatic tail in previous studies, there is no direct evidence showing the cyclosporin A-induced cells share the source in this manuscript. Without any proper lineage tracing work, the origin of those cyclosporin A-induced cells cannot be concluded.

3. It is interesting to see an increase of beta cells in the primary islet after cyclosporin A treatment (Supplemental Fig 2B). However, it remains unclear if their formation shares the same mechanism with the newly formed beta cells in the pancreatic tail.

4. The conclusion of the effect of cyclosporin A on the endocrine progenitors (Line 175) is not convincing because the data cannot distinguish the endocrine progenitors from the insulin-expressing cells. Indeed, Figure 2E shows that neurod1+ cells are fewer than ins+ cells (Figure 2D) in the pancreatic tail at 10 dpt, suggesting that all or at least the majority of neurod1+ cells are already ins+.

5. Figure 5D shows a significant loss of nkx6.1+ cells in the combined treatment group but there is no direct evidence showing this was a result of differentiation as the authors suggested. This cell loss also outnumbered the increase in ins+ cells (Figure 4D). The cell fates of these lost cells are still undetermined, and the authors did not demonstrate if apoptosis could be a reason of the cell loss.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment

This work presents some valuable information regarding the molecular mechanisms controlling the regeneration of pancreatic beta cells following induced cell ablation. However, the study lacks the critical lineage tracing result to support the conclusion about the origin of the regenerated beta cells. The results of the pharmacological manipulation of CaN signaling are also incomplete. In particular, these manipulation are not cell-specific, making it difficult to interpret and thus genetic approach is recommended.

Public Reviews:

Reviewer #1 (Public Review):

Induction of beta cell regeneration is a promising approach for the treatment of diabetes. In this study, Massoz et.al., identified calcineurin (CaN) as a new potential modulator of beta cell regeneration by using zebrafish as model. They also showed that calcineurin (CaN) works together with Notch signaling calcineurin (CaN) to promote the beta cell regeneration. Overall, the paper is well organized, and technically sound. However, some evidence seems weak to get the conclusion.

Reviewer #2 (Public Review):

This work started with transcriptomic profiling of ductal cells to identify the upregulation of calcineurin in the zebrafish after beta-cell ablation. By suppressing calcineurin with its chemical inhibitor cyclosporin A and expressing a constitutively active form of calcineurin ubiquitously or specifically in ductal cells, the authors found that inhibited calcineurin activity promoted beta-cell regeneration transiently while ectopic calcineurin activity hindered beta-cell regeneration in the pancreatic tail. They also showed similar effects in the basal state but only when it was within a particular permissive window of Notch activity. To further investigate the roles of calcineurin in the ductal cells, the authors demonstrated that calcineurin inhibition additionally induced the proliferation of the ductal cells in the regenerative context or under a limited level of Notch activity. Interestingly, the enhanced proliferation was followed by a depletion of ductal cells, suggesting that calcineurin inhibition would exhaust the ductal cells. Based on the data, the authors proposed a very attractive and intriguing model of the role of calcineurin in maintaining the balance of the progenitor proliferation and the endocrine differentiation. However, the conclusions of this paper are only partially supported by the data as some evidence from the data remains suggestive.

(1) In the transcriptomic profiling, genes differentially regulated in the ablated adults could be solely due to the chemical effects of metronidazole instead of the beta-cell ablation. A control group without ins:NTR-mCherry but treated with metronidazole is necessary to exclude the side effects of metronidazole.

We believe that it is unlikely that the differential regulation observed is due to metronidazole rather than the beta cell loss. This experimental strategy as proven successful in well-published studies to identify regulators of beta cell regeneration in the zebrafish larvae. Importantly, the candidates identified in these studies were subsequently functionally validated in mammalian models (Lu et al. 2016, Karampelias 2021). Moreover, in our study, we also used another chemical compound, the nifurpirinol (Bergemann et al., 2018), to ablate the beta cells. Regardless of whether we employed metronidazole or nifurpirinol for beta cell ablation, our results consistently indicate a notable involvement of calcineurin. Of note, the nifurpirinol molecule is commonly used in fishkeeping without toxicity reported on the global health of the fish.

(2) Although it has been shown that the pancreatic duct is a major source of the secondary islets in the pancreatic tail in previous studies, there is no direct evidence showing the cyclosporin A-induced cells share the source in this manuscript. Without any proper lineage tracing work, the origin of those cyclosporin A-induced cells cannot be concluded.

Our experimental setting is similar to the one described in Ninov et al. 2013, where lineage tracing experiments demonstrate an increase of beta cell formation in the pancreatic tail that originate from the pancreatic ducts. In our study, we performed the same experiment with the addition of CsA and showed more ductal cell proliferation (Figure 5G) followed by a 19% increase of beta cell regeneration compared to nonregenerative conditions (Figure 2B). It is unlikely that the additional 19% of regenerated beta cells under CaN inhibition come from another source than the 68% first.

On the other hand, the acinar cells cannot be consider as another source of regenerated beta cell as they are not able to form beta cells unless they are artificially reprogrammed (Maddison et al., 2012). Therefore the only other potential source of regenerated beta cell is the endocrine compartment. However at the stage where we performed beta cell ablation, there are no endocrine cell in the pancreatic tail. Moreover, there are no evidence that secondary islets could come from the principal islet, they are tightly associated with the ducts and differentiate form ductal cell (Mi et al., 2023).

Importantly, we demonstrated that overexpression of CaN specifically in the pancreatic ducts prevents beta cell regeneration. CaN effect is therefore intrinsic to the ducts. Moreover, we showed that CsA increase beta cells formation when Notch signalling is repressed. Given that Notch signalling is known to act on the ductal cell population, this strongly suggests again that CsA exacerbate beta cells formation from the ducts.

All of these compelling evidences strongly support the notion that the cyclosporininduced beta cells originate from the ductal cells.

(3) It is interesting to see an increase of beta cells in the primary islet after cyclosporin A treatment (Supplemental Fig 2B). However, it remains unclear if their formation shares the same mechanism with the newly formed beta cells in the pancreatic tail.

There are indeed several source of beta cell regeneration in the primary islet. However, a recent study showed that the contribution of alpha cell to regeneration is minor and the main contributors are ductal and sst1.1 cells (Mi et al., 2023). In our previous publication, we indeed showed that a major source of beta cell in the principal islet is the delta 1.1 cell population. Those sst1.1 cells begin to express insulin and therefore are named ‘bihormonal’ (Carril et al., 2022). We tested if this population is impacted by CsA treatment and we showed below that CsA does not affect bi-hormonal cell formation (Figure 2D supplemental). These new results suggest that the CsA mediated increase of beta cells in the principal islet arise from the ductal cells as observed in the tail. These results were added in the manuscript as Figure 2D supplemental.

Tg (sst1.

1:GFP); Tg (ins:NTR*-mCherry) larvae were treated at 3dpf with NFP 4µM to induce beta cell ablation. Then larvae were treated with CsA 1µM from 4 to 6 dpf (or ctl with DMSO); prior fixation and analysis of bi-hormonal cells in the principal islet at 6dpf.

(4) The conclusion of the effect of cyclosporin A on the endocrine progenitors (Line 175) is not convincing because the data cannot distinguish the endocrine progenitors from the insulin-expressing cells. Indeed, Figure 2E shows that neurod1+ cells are fewer than ins+ cells (Figure 2D) in the pancreatic tail at 10 dpt, suggesting that all or at least the majority of neurod1+ cells are already ins+.

The neurod1+ cells population indeed included both endocrine progenitor cells and differentiated endocrine cells. However, we would like to point out that the timing of the analysis is essential to reach our conclusion. When we treat with CsA, we show an increase of neurod1+ cells already at 4dpt. At this time point, no hormone- producing cell can yet be detected (Figure 2E). Those additional neurod1+ cell are therefore endocrine progenitors and not beta cells. This result shows that CaN inhibition induces pro-endocrine cell formation in regenerative conditions.

At 10dpt, the neurod1+ cells population includes beta cells as well as endocrine progenitor cell. We agree that the way the data are presented in figure 2D and 2E can be confusing. Those 2 figures come form 2 separated experiments, the number of beta cell in figure 2D can therefore not be compared to the number of Neurod1+ cell in figure 2E. Indeed, from one experiment to another the efficiency and rate of regeneration can vary, independently of calcineurin. To clarify, we added the number of beta cells regenerated in the experiment of figure 2E (see Author response image 2 in red). As you can see in this experiment, regeneration was a bit slower than usual.

Tg (neurod1:GFP); Tg (ins:NTR*-mCherry) larvae were treated at 3dpf with NFP 4µM to induce beta cell ablation.

Then larvae were treated with CsA 1µM from 4 to 6 dpf (or ctl with DMSO); prior fixation and analysis of GFP+ cells (in grey, pink, dark grey and green), and mCherry+ cells for the condition ablated + CsA in red from 2 to 10 dpf.

(5) Figure 5D shows a significant loss of nkx6.1+ cells in the combined treatment group but there is no direct evidence showing this was a result of differentiation as the authors suggested. This cell loss also outnumbered the increase in ins+ cells (Figure 4D). The cell fates of these lost cells are still undetermined, and the authors did not demonstrate if apoptosis could be a reason of the cell loss.

Firstly, as you can notice on the graphs, we encountered a very high variability between individuals within the same condition. We decided to show this variability by presenting the raw data. This high variability could partially explain the differences that you underline. Moreover, we would like to point out that independently of CaN inhibition the progenitor loss (nkx6.1+ cell) outnumber the gain of beta cells. Indeed, in average there is a loss of 29% (41 GFP+) of the nkx6.1+ cells and a gain of only 6 beta cells after Notch inhibitory treatment. The other progenitors cells being differentiated into other endocrine cell types (pro-endocrine, alpha, delta). In the combined treatment (Notch and CaN inhibitors), we decreased the number of progenitors cell by 50%, i.e 21% (20 cells) more than without CaN inhibitor. However, we increased the number of regenerated beta cells by two fold (6 cell to 12 cells). In brief, the important progenitors cell loss could be explained by precocious differentiation in the pro-endocrine and endocrine cells type. It is therefore normal than the number of beta cells regenerated do not match the progenitors cell number loss and this in presence or absence of CaN inhibition.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Major concerns:

(1) The evidence to indicate the proliferating ductal cell differentiate into beta cell is weak. They should use linkage tracing, or other marker genes immunostaining to confirm that.

The experiment from the Figure 5 A-D is a short term tracing experiment and should have been presented as such in the manuscript. After LY411575 (Notch inhibitor) and CsA treatments at 3dpf, we exposed the larvae to EdU at 4dpf during 8 hours (Figure 5A). We showed that EdU is incorporated in dividing ductal cells at 4dpf (Figure 5C) ant that 2 days later there are newly form beta cells that are EdU+.(see Author response image 3) To reinforce our conclusion, the image below will be added to the manuscript.

Tg (nkx6.

1:GFP); Tg (ins:NTR*-mCherry) larvae were treated at 3dpf with both CsA 1µM and LY411575 5µM. At 4dpf, the larvae were exposed to EdU 4mM during 8 hours, before analysis at 6 dpf.

(2) To inhibition of CaN and Notch pathway, they just used the pharmacological approaches, genetical approaches should be used to get stronger evidence.

We employed two distinct inhibitors specifically targeting calcineurin (CsA and FK506) for CaN inhibition. While these inhibitors have distinct chemical structures and potential non-specific effects, they both yield the same result of increased beta cell formation under Notch repression (see Figure 4D and Figure 4B in the supplementary data). This convergence of outcomes strongly suggests that the observed effect is primarily attributable to the specific inhibition of calcineurin.

Furthermore, we complemented our inhibitor-based approach with a genetic strategy involving CaN overexpression (see Figure 3). Notably, the overactivation of CaN resulted in a reduction of beta cell regeneration. Given that this genetic approach generated an effect contrary to that achieved with the inhibitors, it provides robust support for our model, which postulates that calcineurin plays a critical role in the regulation of beta cell regeneration (see Figure 3, panels C-E).

As for Notch inhibition, previous published data from our laboratory compared the effects of Notch inhibitor (LY411575) and genetic approaches (mib mutant and transgenic line) on pro-endocrine cell (ascl1b+) and ductal cell (nkx6.1+) formation. This study showed that both Notch inhibitor (LY411575) and Notch repression using genetic approaches recapitulate the same effect: an induction of pro-endocrine cells formation. The specificity of this inhibitor being validated (Ghaye et al., 2015), we did not consider the need of a genetic approach.

(3) The most enriched pathways among the up-regulated genes were DNA replication and cell cycle, which suggested that these genes are more important for the duct cell proliferation, how is Calcineurin related to these pathways, such as regulating the genes important for proliferation?

The transcriptomic data presented in this manuscript suggest that the ductal cells undergo a strong proliferative response after beta cell ablation. This is in accordance with our experimental data showing activation of ductal proliferation after beta cell ablation (Ghaye at al., 2015) and data from this manuscript (Figure 1 I-J).

Calcineurin is a well-known regulator of the cell cycle, and can either promote or repress the cell cycle depending on the cell type. For example, stressing the cell provokes an entry of calcium and subsequently a CaN activation which result in cell cycle arrest (Leech et al. 2020). Nevertheless, depending the cell type, CaN can be either necessary or deleterious to cell proliferation (Goshima et al. 2019; Masaki and Shimada 2022). The intriguing dual role of CaN in cell cycle is well illustrated in β cell regeneration. While CaN should be repressed to enable ductal progenitor amplification and subsequent endocrine differentiation, CaN is then necessary for β cell function and for their replication (Dai et al. 2017; Heit et al. 2006). Moreover, CaN is related to cellular senescence and CaN function is important for proper fin regeneration in zebrafish.

(4) It is hard to understand why they pick up the pathway of cellular senescence signature for the duct cell progenitor neogenesis? Moreover, among these senescence genes, many genes are cell cycle regulators.

In response to beta cell ablation, the ductal cells undergo a strong proliferative response, as shown in our previous data (Ghaye 2015). It was therefore not surprising that many differentially expressed genes are cell cycle regulators. On the other hand, the cellular senescence signature was surprising. Indeed, senescence is usually associated with cell cycle arrest and aging. However, recent studies showed that cellular senescence is required for proper development and regeneration. We therefore wanted to investigate this pathway and more particularly the function of calcineurin, which can either promote or repress the cell cycle in different cell types (see comment above).

(5) The RNA-seq data obtained from adult fish, while the authors use larvae to explore the CaN functions, it may have different conclusion using adult fish. Moreover, it is unclear whether the CaN increased when the beta cell ablated in young larvae.

We decided to first perform functional experiment in the larvae as this model unable the quantification of beta cell regeneration from the ducts in the pancreatic tail. However, to validate our results in non-developmental stages, we perform experiments in juveniles (2 months old) and adults. CsA treatments in juveniles zebrafish recapitulated the same results that in larvae (Figure 2B and Figure 6A-C). Moreover, we showed that CaN overactivation delayed glycemia recovery after ablation adults (Figure 6D-E), which is in accordance with an impaired regeneration. Altogether, these results strongly suggest that CaN act as regulator of beta cell regeneration both in the juvenile/adult and larval stages.

Concerning the expression of CaN in the zebrafish larvae, we tried to detect the level of CaN in the different experimental conditions by in situ hybridization. However, we were not able to detect it using this technique. We also tried immunostaining with antiphospho-nfact3 ser165 polyclonal antibody (Invitrogen) but this antibody does not seem to work in zebrafish. Finally, we tried to sort ductal cell at larval stage to perform a transcriptomic analysis but we were unable to collect enough ductal cells to proceed further. Indeed our staining experiment showed that there are only around 150 ductal cells (nkx6.1+, Figure 5D) at this stage.

(6) The beta cell regeneration in the young larvae usually recovers within ~ 5 days in principle islet. Please also show the beta cell number (PI) during the beta cell recovery after ablation.

We did show beta cell regeneration in the principal islet in Figure 2A-B supplemental. While new beta cells appears quickly in this islet (Carril, Massoz, Dupont et al., 2023), the principal islet has not yet fully recover at 5dpt.

(7) Since the studies did not show the CaN level in Fig.3, it is hard to know that the CaN is exactly expressed.

In the figure 3B, using Tg(hsp70:GFP-CaNCA), it is indeed not possible to see CaN expression at 10 dpt as the heat shocks induce only transiently CaNCA overexpression. However, the transient expression was detected in live shortly after the heat shocks. On the other hand, with the transgenic line Tg(UAS:GFP-CaNCA); Tg(cftr:Gal4), in which GFPCaNCA is continuously expressed allowing us to show CaNCA expression in the pancreatic ducts (Figure 3).

(8) In Fig.6 D and 6E, did these drug treatments change the glucose level in nonablated fish?

As you can see below, the CaN inhibitor, CsA does not affect the glycemia of the fish in non-regenerative conditions.

Glycemia of non-ablated fish, 3 days after drug treatment.

(9) The logic of writing in Results is very hard to understand.

We proofed read the paper in an effort to clarify it.

Minor concerns,

(1) Make a scheme for ablation and RNA-seq, and indicate the age of the fish used in Fig. 1.

We added the scheme in Figure 1 supplemental.

(2) In Fig. 1G, two arrows indicated mCherry+ cells is hard to see in the non-ablated fish.

One arrow was indeed mislocated, we moved the arrow and try to improve the intensity of red. However, the only cells are indeed small and can be difficult to see.

(3) In Fig.6, it is hard to know that the arrows indicated islets are small islets (up to 5 cells), how they compared with big islets and defined as small islet. Moreover, some of these islets are almost invisible.

We now show a close up of a portion of the pancreatic tail and show the beta cells with arrows only in this picture, to enhance clarity.

Reviewer #2 (Recommendations For The Authors):

(1) This manuscript needs more proofreading and polishing to increase its readability.

We proofread the manuscript and change some paragraph for more clarity.

(2) The extensive use of words like "modulate" or "regulate" sometimes makes the text ambiguous as the effect is not stated directly and clearly.

We re-wrote some parts of the text and try to avoid using “regulate” as often.

However, as we used both repression and over-activation of CaN, we still use words as regulate to stipulate general conclusions on the function of CaN.

(3) The list of individual differentially regulated genes after the beta-cell ablation in the RNAseq seems missing. This list could be interesting and helpful for other researchers. We added it.

(4) In Figure 1D, "modulated" genes are shown but were they all upregulated like those in Figure 1A? The modulation should be indicated more clearly (e.g. up- or down-regulated) in the figure. The authors can use different colours to illustrate that.

Done.

(5) Is Figure 2D showing the same data extracted from Figure 2B? Does Figure 2D add any information to the data?

No, it does not add data. We actually add the Figure 2D for a better visualisation of the increase at 10dpt.

(6) In the y-axis of Figure 3E, it should be "mCherry".

It already is. We did check all the axis again to be sure it is correct.

(7) Line 219, "Figure 4E supplemental" instead of "Figure 4D supplemental"

Done.

(8) Line 266, "ablated juveniles" instead of "ablated larvae"

Done. Thank you for noticing these mistakes.

(9) In Figure 6A, many mCherry+ cells are hardly visible and there are some greyish white signals in the images that are supposed to show the mCherry channel only. What are those grey signals?

There is no channel showing grey on the picture, I improved the overall quality of this pictures and show close up to improve the figure.

(10) In Figure 6D and 6E, CaNCA overexpression had a significant effect on the glycemia. But did the overexpression affect the beta cell formation or regeneration?

We showed that CaNCA overexpression did not affect beta cell formation in absence of regeneration in the larvae (Figure 3E). Moreover, it does not affect the glycemia of the fish in non-regenerative conditions (Author response image 5). As for regenerative conditions, CaN overexpression decreased the regeneration in the larvae (Figure 3E).

Glycemia of Tg(UAS:GFP-CaNCA); Tg(cftr:Gal4) fish, overexpressing CaNCA, compared to controls fish, in non-regenerative conditions.

(11) The role of calcineurin seems transient (e.g. Figure 2B and 4E) and does not play a significant role in long term. It would be interesting to see if long-term/repeated treatments of calcineurin inhibitors and overexpression/knockout of important members of calcineurin signaling would affect the pool of progenitors in long term.

We were also interested in the consequences of CaN overexpression on the long term. Our overexpression tool Tg(UAS:CaNCA) allow to address this question, as CaN is overexpress permanently. We assessed the structure of the ducts and the number of beta cells in transgenic larvae and did not see any defects of the ducts whether in regenerative context or not. On the other hand, we showed in this manuscript that CaN effect is specific to regenerative conditions. As a consequence, it is not likely that repeated treatments long after the ablation would continue to affect beta cell formation and the progenitors pool.

Version 2

Abstract

Abstract

Stimulation of pancreatic beta cell regeneration could be a therapeutic lead to treat diabetes. Unlike humans, the zebrafish can efficiently regenerate beta cells, notably from ductal pancreatic progenitors. To gain insight into the molecular pathways involved in this process, we established the transcriptomic profile of the ductal cells after beta cell ablation in the adult zebrafish. These data highlighted the protein phosphatase calcineurin as a new potential modulator of beta cell regeneration. We showed that calcineurin overexpression abolished the regenerative response, leading to glycemia dysregulation. On the opposite, calcineurin inhibition increased ductal cell proliferation and subsequent beta cell regeneration. Interestingly, the enhanced proliferation of the progenitors was paradoxically coupled with their exhaustion. This suggests that the proliferating progenitors are next entering in differentiation. Calcineurin appears as a guardian which prevents an excessive progenitor proliferation to preserve the pool of progenitors. Altogether, our findings reveal calcineurin as a key player in the balance between proliferation and differentiation to enable a proper beta cell regeneration.

Introduction

Blood glucose homeostasis is tightly controlled by pancreatic endocrine cells. Insulin-producing beta cells work in close association with alpha cells, which secrete glucagon, and delta cells secreting somatostatin, to maintain normal glycemia. The insulin plays a critical role in this process as it is the only hormone able to lower the glycemia. Beta cell loss is a hallmark of type 1 and of late stages of type 2 diabetes, leading to chronic hyperglycemia. Beta cell destruction is largely irreversible in human and in mammals, making the disease incurable nowadays. Nevertheless, studies in diabetic mice models uncovered promising evidences of slight recovery of beta cells via regeneration. For example, new beta cells can arise from the replication of remaining beta cells [1]. Different models of pancreas injuries revealed the plasticity of mammalian pancreatic cells. Actually, differentiated endocrine cells such as alpha cells[2] and delta cells[3] can reprogram and convert into insulin-producing cells, a process that is age dependent. Another possible source of beta cells could be de novo formation from pancreatic progenitors residing in the ductal compartment of the adult pancreas [4]. Lineage tracing experiments could not clearly highlight new beta cells arising from the ductal cells in the adult [4]–[6]. However, expression of the pro-endocrine marker Neurog3 was detected in the ducts in different mouse models of pancreas regeneration [4], [7], [8]. In addition, a rare population of ductal cells expressing Neurog3 has been reported to contribute to beta cell neogenesis during diabetes [9]. A recent study of single cell RNA sequencing reveal a sub-population of human ductal cells which are able to give rise to all the pancreatic cell types, including beta cells, following implantation in mice [10]. Altogether, these findings suggest that mammalian pancreatic ducts possess the intrinsic capacity to (re)generate beta cells even though this process is poorly efficient and slow, especially in adults. In contrast to mammals, zebrafish possess remarkable capacity of regeneration, independently of its age [11]–[16]. This model can therefore be exploited to identify and characterize regenerative mechanisms and ultimately induce regeneration in mammals. Based on regenerative mechanisms identified in zebrafish, several studies succeed to improve regeneration in mammals, underlying the possibility of translation from zebrafish to mammals ([17]–[21] and [22], for review).

In zebrafish, the nitroreductase (NTR)/nitroaromatic prodrug system is widely used. In this technique, the combination of cell type-specific NTR expression and nitroaromatic prodrug exposure allow for controlled and targeted cell ablation [23], [24]. When NTR is expressed under the control of the insulin (ins) promoter (Tg(ins:NTR-mCherry) [25]), beta cell destruction is complete within 3 days following the treatment in adult fish, which correlate with a peak of hyperglycemia [26], [27]. The regeneration of beta cells upon ablation is spontaneous and fast and the glycemia is normalized within two weeks [13], [29]. Similar to mice, new beta cells can arise through proliferation of surviving beta cells [28] as well as through the contribution of alpha cells [29], [30] and delta cells[31], [32], underscoring the overall conservation of these processes from zebrafish to mammals. Nevertheless, unlike mammals, the presence of pancreatic progenitors in the ducts is well established in both larval [33] and adult zebrafish [26], [27]. Lineage tracing experiments pointed out that ductal cells and centroacinar cells (CACs) give rise to new beta cells [26], [33]. In adults, duct-derived beta cells start to be detected between 7 and 10 days following beta cell ablation [26], [27].

Notch signaling is a key player regulating the differentiation of duct-associated progenitors into endocrine cells. Larval duct cells as well as adult CACs display strong Notch activity [26], [34]. This signaling pathway has a central role in beta cell genesis during both development [35] and regeneration [33] by repressing endocrine differentiation. In zebrafish, different levels of Notch activity determine the behavior of the pancreatic progenitors. While a high level of Notch activity maintains cells in quiescence, a moderate level induces the entry in the cell cycle and proliferation whereas a low level drives endocrine differentiation of the progenitor cells [35]. A steep decrease of Notch activity pushes the progenitors to differentiate prematurely, bypassing the amplification step and leading to their depletion [35]. Repression of the Notch signaling by mTor, activated by glucose and nutrients, hence promotes beta cell formation and regeneration from ductal progenitors [33]. While Notch and mTor signaling are crucial for this process, there is still a need to establish a global view of the molecular mechanisms regulating beta cell regeneration.

To identify early events regulating ductal-derived beta cell regeneration, we determined the transcriptomic signature of ductal cells from adult zebrafish following beta cell destruction. Our data highlighted an upregulation of the Calcineurin (CaN) pathway. To elucidate CaN function in beta cell regeneration, we both repressed and activated CaN pathway. We showed that CaN regulates beta cell neogenesis in the ducts during regeneration, by acting on progenitor proliferation. Together, our findings underline that CaN fine tunes the balance between progenitor proliferation and beta cell differentiation to guarantee proper regeneration.

Results

Transcriptomic profiling of ductal cells after beta cell destruction highlights regulation of calcineurin pathway

To gain a better understanding of the molecular mechanisms underlying the regeneration of beta cells from the ducts, we determined the transcriptional landscape of ductal cells by RNA-sequencing after beta cell ablation in the adult zebrafish. To selectively ablate the beta cells, we used the Tg(ins:NTR-mCherry) transgenic fish. The ductal cells were labelled thanks to the Tg(nkx6.1:GFP) reporter line [27] in which GFP marks the ductal tree and associated multipotent pancreatic progenitors [27]. More precisely, 3 to 4 fishes were treated with the pro-drug MTZ at 10mM overnight, to induce ablation of beta cells or with DMSO for the non-ablated controls. Ablation was confirmed by blood glucose measurement before collection of the pancreas. To capture the early events triggered by the destruction of beta cells, we generated the transcriptome of the ducts 3 days post ablation treatment (dpt), i.e. before beta cell neogenesis (Figure 1A supplemental). Differential gene expression analysis revealed that 1866 genes are up-regulated and 1515 genes down-regulated in the ductal cell of fish treated with MTZ compared to control (pAdj < 0.05). According to Gene Ontology analysis, the most enriched pathways among the up-regulated genes were DNA replication and cell cycle (Figure 1A). This further corroborates our previous findings and those of others, regarding the activation of duct-associated progenitors’ proliferation in response to beta cell ablation [26], [27]. As expected, the Notch pathway was enriched in the down-regulated genes [33], [35] (Figure 1B). On a contrasting note, in a context marked by a robust proliferative response, we found it intriguing to observe an enrichment in the cellular senescence signature (Figure 1A). Subsequently, we investigated deeper into the genes associated with this specific signature. We found among them several components of the calcineurin signaling pathway such as nfatc3b, ppp3ccb (the catalytic subunit of Calcineurin), itpr2 and calm3b (Figure 1C). In addition to these genes related to the cellular senescence signature, our transcriptomic studies revealed the modulation of other genes from the CaN canonical pathway (Figure 1D), underlying its potential role in beta cell regeneration. Calcineurin (CaN) is a highly conserved calcium/calmodulin dependent Ser/Thr phosphatase, involved in numerous biological process including fin regeneration and beta cell function [36]–[39]. This prompted us to investigate the role of CaN in beta cell regeneration.

Previous RNAseq data performed in our laboratory indicate that CaN (ppp3cca/b) and NFATc3 (nfatc3a/b) are mainly express in endocrine cells [40] (Figure 1B supplemental), which is in accordance with the role of CaN/NFAT signalling in beta cells [41]. CaN genes (ppp3cca/b) as well as nfatc3b, are express at lower levels in the ducts at basal state but their expression is induced in response to beta cell destruction (Figure 1C supplemental).

Calcineurin activity regulates the ductal regenerative response

To evaluate the role of CaN in beta cell regeneration, more specifically derived from ductal progenitors, we have chosen to use young larvae, where regenerated beta cells in the pancreatic tail arise exclusively from the ducts [33]. In response to beta cells ablation, the intrapancreatic ducts undergo a ductal regenerative response whereby differentiation toward the endocrine fate is increased [33]. We first determined the rate of beta cell neogenesis from the ducts in response to a single acute ablation of beta cells as we performed in adults. We treated Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae with nifurpirinol (NFP) from 3 to 4 dpf and mCherry+ beta cells were quantified in the GFP+ ducts in the pancreatic tail at several time points: 4, 7, 10 and 14 dpt (Figure 1E). Duct-associated beta cells started to be detected in non-ablated larvae between 7 and 10 dpt (Figure 1F) and the number of beta cells slowly increased until 14 dpt (Figure 1F, 1G). In ablated larvae, the increase became more pronounced from 10 dpt onwards (Figure 1F, 1G), indicating faster endocrine differentiation. This experiment establishes that the ductal regenerative response is detectable between 10 and 14 days after the beta cell ablation, performed at 3 dpf. We next wanted to determine if ductal cell proliferation is activated in response to beta cell destruction in larvae as in adult fish. We exposed Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae to EdU the second day following ablation (Figure 1H). In the ablated larvae, the proportion of GFP+ ductal cells EdU+ (in S-phase) was higher compared to non-ablated larvae (Figure 1I, 1J). This result shows that acute beta cell ablation in larvae rapidly activates ductal cell proliferation, as previously reported in adult zebrafish.

As our transcriptomic data from adult zebrafish revealed modulation of the CaN pathway at 3 dpt (Figure 1D), we treated Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae from 1 to 3 dpt after beta cell ablation, with a CaN inhibitor, the Cyclosporin A (CsA) [39] (Figure 2A). The number of newly formed beta cells in the tail was monitored from 4 to 14 dpt. CsA enhanced beta cell formation at 10 dpt (Figure 2B-D). However, this effect appears to be transient since no discernible difference was observed between the control and CsA-treated larvae in regeneration at the latest time point (14dpt), suggesting an acceleration of the regenerative response (Figure 2B, Figure 2A supplemental). Interestingly, CsA did not affect beta cell differentiation in non-ablated larvae, indicating that CsA only acts in a regenerative context (Figure 2B-D). Of note, CsA increased as well the number of regenerated beta cells in the principal islet (Figure 2B, 2C supplemental). We and others, previously showed that the other major source of beta cell regeneration, besides the ducts, are the bi-hormonal sst1.1+/ins+ cells [31], [32], [42]. However, CsA did not affect bi-hormonal cell formation (Figure 2D supplemental), suggesting that the additional cells in the principal islet could also originate from the ducts.

Transcriptomic profiling of ductal cells during beta cell regeneration and validation in larvae

A-B) Enrichment ratio of selected non-redundant signatures of KEGG pathways overrepresented in ductal cells after beta cells ablation (UP – A and DOWN-B) compared to ductal cells without beta cells ablation. Gene Ontology (GO) terms were identified using ORA analysis by WebGestalt (Liao et al., 2019) using the list of DE genes provided by DESeq. The light color for Notch pathway means p-value = 0.11.

C) List of genes associated with the signature of cellular senescence from A-B. Genes related to CaN pathway are in bold.

D) Calcineurin canonical pathway with up-regulated genes in green and down-regulated genes in red in transcriptomic data from A-B.

E) Experimental design for regeneration test in larvae. Briefly, after nifurpirinol treatment from 3 to 4dpf, larvae were fixed and analyzed at 4-7-10 and 14 days post treatment (dpt).

F) Graph representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) at 0-4-7-10 and 14 dpt. The gray spheres represent non-ablated conditions and the pink triangles the ablated condition. Data are presented as mean values ± SEM. One-way ANOVA test with Tukey’s multiple comparison test, ****p-value<0.0005.

G) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 14dpt. 3D projection (stack) of one non-ablated and one ablated representative samples. The principal islet (PI) and the pancreatic tail are shown. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 100µM.

H) Experimental design for EdU assay in larvae. After NFP treatment for 3 to 4dpf, larvae were exposed to EdU at 2dpt before fixation for analysis.

I) Whole mount fluorescent immunohistochemistry (GFP and EdU) of the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 2dpt. 3D projection (stack) of one non-ablated and one ablated representative samples. Arrows point out GFP+ duct cells EdU+ in the pancreatic tail. Scale 50µM.

J) Barplot representing the percentage of GFP+ ductal cells which incorporated EdU+ in non-ablated and ablated conditions. Data are presented as mean values ± SD. T-test. **p-value<0.005.

Calcineurin inhibition with CsA increases the ductal regenerative response

A) Experimental design for regeneration test in larvae with CsA treatment. Briefly, after nifurpirinol treatment from 3 to 4dpf, larvae were treated with CsA from 1 to 3dpt and fixed and analyzed at 4-7-10 and 14 days post treatment (dpt).

B) Graph representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) at 0-4-7-10 and 14 dpt. The gray spheres represent non-ablated condition; the pink triangles represent the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SEM. Two-way ANOVA test with Sidak’s multiple comparisons test, *p-value<0.05.

C) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 10dpt. 3D projection (stack) of non-ablated and ablated larvae treated with DMSO or CsA representative samples. The principal islet (PI) and the pancreatic tail are shown. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 100µM.

D) Barplot representing the number of number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 10dpt. The gray spheres represent non-ablated condition; the pink triangles represent the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SD. Two-way ANOVA with Tukey’s multiple comparison test, *p-value<0.05.

E) Graph representing the mean number of GFP+ neurod1+ cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(neurod1:GFP) at 0-4-7 and 10 dpt. The gray spheres represent non-ablated condition; the pink triangles represent the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SEM. Two-way ANOVA test with Sidak’s multiple comparisons test, **p-value<0.005; *** p-value<0.0005; **** p-value<0.00005.

F) Graph representing the mean number of GFP+ neurod1 EdU+ cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(neurod1:GFP) at 0-4-7 and 10 dpt. The gray spheres represent non-ablated condition; the pink triangles represent the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SEM.

We next tested the effect of CsA on endocrine progenitors in a regenerative context. We induced regeneration in Tg(neurod1:GFP) larvae where the GFP is expressed in both endocrine progenitors and mature endocrine cells. We first assessed generation of GFP cells at different time points (Figure 2E supplemental) and showed that CsA induced an increase of neurod1+ cells from 4 dpt. The increase was still detectable at least until 10dpt (Figure 2E, Figure 2F supplemental). To determine if these additional cells result from their own proliferation, we performed a pulse of EdU just before analysis (Figure 2E supplemental). We observed that CsA did not affect the neurod1+ cell proliferation rate, which is very low at these stages (Figure 2F). As a consequence, the effect of CsA cannot be explained by endocrine cell proliferation but rather by neogenesis from progenitors. As CsA affects pro-endocrine cells formation, we next wondered if the increased cell formation induced by CsA is specific to beta cells. Treatment with CsA was performed as previously and delta1.1 and alpha cells were detected by immunofluorescence. Interestingly, CsA did not affect alpha nor delta1.1 cells neogenesis in response to beta cell ablation (Figure 2G-H supplemental). Overall, these experiments showed that CsA affects specifically the beta cells and their endocrine progenitors.

CaN over-activation abolishes the regenerative response

We then wondered if an opposite regulation of CaN i.e. its activation impacts as well the regenerative response. To that end, we generated a transgenic line Tg(hsp70:GFP-P2A-ppp3ccCA) that allows ubiquitous expression of a constitutively active form of CaNCA (ppp3ccCA) upon heat shocks. Beta cell ablation was triggered in Tg(hsp70:GFP-P2A-ppp3ccCA); Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae from 3 to 4 dpf and CaNCA expression was induced by four successive heat-shocks from 1 to 3 dpt (Figure 3A). The overexpression of CaNCA after ablation impaired the regenerative response at 14 dpt (Figure 3B-C). Similar results were obtained with Tg(UAS:GFP-P2A-ppp3ccCA); Tg(cftr:gal4)[45]; Tg(ins:NTR-P2A-mCherry) larvae in which CaNCA is continuously and specifically overexpressed in the ducts within the pancreas (Figure 3D-E). Importantly, the structure of the ducts was similar in CaNCA overexpressing larvae compared to Tg(nkx6.1:GFP) controls (Figure 3D), suggesting that the suppression of the regenerative response in CaNCA-overexpressing larvae was not due to morphogenetic defects during ductal growth. This result shows that CaN acts directly in the ducts to decrease beta cell regeneration while it is not necessary for normal beta cell differentiation.

Transgenic mediated overexpression of calcineurin abolish the ductal regenerative response

A) Experimental design for regeneration test in larvae with heat-shocks. Briefly, after nifurpirinol treatment from 3 to 4dpf, four heat shock were performed from 1 to 3dpt and larvae were fixed and analyzed at 14dpt.

B) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(hsp70:CaNCA); Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 14dpt. 3D projection (stack) of one non-ablated and one ablated with or without heat-shock representative samples. The principal islet (PI) and the pancreatic tail are showed. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 100µM.

C) Barplot representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(hsp70:CaNCA); Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 14dpt. The gray spheres represent non-ablated condition; the pink triangles represent the ablated condition; the black squares heat-shock condition and inverted green triangles ablated + heat-shock condition. Data are presented as mean values ± SD. Two-way ANOVA with Tukey’s multiple comparisons test, *p-value<0.05, *** p-value<0.0005, ns = non-significant.

D) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of larvae at 14dpt. 3D projection (stack) of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) one non-ablated and one ablated representative control samples and Tg(UAS:CaNCA); Tg(ins:NTR-P2A-mCherry); Tg(cftr:gal4) one non-ablated and one ablated representative samples. The principal islet (PI) and the pancreatic tail are showed. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 100µM.

E) Barplot representing the mean number of mCherry+ beta cells in the pancreatic tail of larvae at 14dpt. The gray spheres represent non-ablated Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) condition; the pink triangles represent the ablated Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) condition; the black squares non-ablated Tg(UAS:CaNCA); Tg(ins:NTR-P2A-mCherry); Tg(cftr:gal4) condition and inverted green triangles ablated Tg(UAS:CaNCA); Tg(ins:NTR-P2A-mCherry); Tg(cftr:gal4) condition. Data are presented as mean values ± SD. Two-way ANOVA with Tukey’s multiple comparison test, *** means p-value<0.0005, **** p-value<0.00005, ns = non-significant.

CaN regulates beta cell differentiation induced by Notch inhibition in absence of regeneration

Our transcriptomic data showed that the Notch pathway is downregulated in ductal cells during beta cell regeneration (Figure 1). As the level of Notch activity determines the behavior of ductal cells [33] from quiescence to proliferation and subsequently to beta cell differentiation, we tested whether calcineurin acts together with the Notch pathway on a common pool of ductal progenitors. To inhibit the Notch pathway, we treated larvae with several concentrations of the gamma-secretase inhibitor LY411575 from 3 to 4dpf in absence of regeneration. The activity of CaN was inhibited by CsA during the same timeframe (Figure 4A). As previously, we used reporter lines for beta and ductal cells Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) and the number of secondary beta cells was analyzed at 6dpf (Figure 4A). As expected, the number of beta cells progressively rose as the concentration of the Notch inhibitor increased (Figure 4B). Combined treatment with CsA fostered the differentiation of beta cells between 1 and 10 µM LY411575 but did not result in further increase at 15 µM LY411575 (Figure 4B), suggesting that CaN is important within a permissive window of Notch activity. Since we overserved the highest synergistic effect at 5µM of LY411575, we used this concentration for the following experiments (Figure 4C-D). It is worth noting that combined treatment of LY411575 (5uM) and another CaN inhibitor, FK506, resulted in the same synergistic increase of beta cell differentiation (Figure 4A-B supplemental), confirming that the effect is due to CaN inhibition. The combined effect of Notch and CaN inhibition is transient as it is not observed at 7dpf anymore (Figure 4E). Therefore, as observed in regenerative conditions (Figure 2B), CaN inhibition accelerates beta cell neogenesis induced by Notch repression.

CaN repression potentializes the effect of Notch inhibition on beta cell formation

A) Experimental design for Notch inhibition test in non-ablated condition. Larvae were treated concomitantly with LY411575 (Notch inhibitor) and CsA from 3 to 4dpf and were fixed and analyzed at 6dpf.

B) Graph representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 6 dpf depending the concentration of LY411575. The blue dots represent LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SEM. Two-way ANOVA test with Sidak’s multiple comparison test, *p-value<0.05, **p-value<0.05, ns = non-significant.

C) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 6dpf. 3D projection (stack) of one control (without any treatment); one CsA-treated; one LY411757-treated and one with both CsA and LY411575 treated larvae. The principal islet (PI) and the pancreatic tail are showed. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 50µM.

D) Barplot representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 6dpf. The black dots represent the control; gray CsA treatment;blue LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SD. Two-way ANOVA with Tukey multiple comparison test, ***p-value<0.0005, ****p-value<0.00005.

E) Graph representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 5-6-7 dpf. The black dots represent the control; gray CsA treatment; blue LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SEM. Two-way ANOVA test with Sidak’s multiple comparison test, *p-value<0.05.

To determine to which extent CaN pathway can modulate Notch induced beta cell neogenesis, we activated CaNCA overexpression in Tg(hsp70:GFP-P2A-ppp3ccCA); Tg(ins:NTR-P2A-mCherry) larvae by an heat-shock at 3 dpf and treated them with LY411575 (Figure 4C supplemental). CaNCA overexpression resulted in a lowered beta cell formation induced by Notch inhibition (Figure 4D supplemental), revealing that CaN activation counterbalanced the effects of Notch inhibition. Using our previous settings of Notch and CaN inhibition, we next wondered if the canonical pathway downstream of CaN was involved in the enhancement of beta cell differentiation. To activate NFAT, a well-known target of CaN, we used CHIR99021 allowing a stabilization of the active form of NFAT (Figure 1D). We found that CHIR99021 rescued the effect of CsA (Figure 4E supplemental), suggesting that CaN inhibition increases beta cell neogenesis at least partially by the regulation of NFAT. Overall, these results reveal that CaN impacts beta cell formation in pro-endocrinogenic context, such as induced by a low level of Notch activity. Moreover, it suggests that both CaN and Notch pathways act on a common pool of ductal progenitors to govern beta cell neogenesis.

CaN controls the proliferation of duct-associated progenitors induced by Notch inhibition

Given that we observed the most significant increase in beta cell formation with CsA when Notch activity was mildly repressed (Figure 4B-D), and since mild Notch activity has been shown to promote progenitor amplification [35], it suggests that CaN acts at this level. To explore this possibility, we exposed briefly Tg(nkx6.1:GFP) larvae to EdU after mild Notch (LY411575 5µM) and CaN inhibition (Figure 5A) and analyzed at 4dpf and 6dpf. As expected, Notch inhibition increased the proportion of proliferating EdU+ GFP+ duct cells at 4dpf (Figure 5B-C) and decreased the amount of ductal progenitors 2 days later (at 6dpf) (Figure 5D), which is concomitant with the increase of beta cell differentiation (Figure 4C-D). Interestingly, the combined inhibition with CsA further increased these proportions, while CaN inhibition alone had no effect (Figure 5B-D). Of note, we detected beta cells EdU+ at 6dpf (Figure 5A supplemental), underlying that these cells originate from the proliferation of ductal progenitors.

CaN repression increases the proportion of ductal proliferating cells

A) Experimental design for EdU assay in Notch test. Larvae were treated concomitantly with LY411575 (Notch inhibitor) and CsA from 3 to 4dpf and then briefly treated with EdU before fixation and analysis at 4 dpf or at 6dpf.

B) Whole mount fluorescent immunohistochemistry (GFP and EdU) of the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 4dpf. 3D projection (stack) of one control (without any treatment), one with CsA only, one with LY411575 only and one with both CsA and LY411757 representative samples. Scale 50µM.

C) Barplot representing the percentage of GFP+ ductal cells which incorporated EdU+ in pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae for the Notch test. The black dots represent the control; gray CsA treatment;blue LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SD. T-test. Two-way ANOVA test with Tukey’s multiple comparisons test, *p-value<0.05; ****p-value<0.00005; ns = non-significant.

D) Barplot representing the number of GFP+ ductal cells which in pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 6dpf for the Notch test. The black dots represent the control; gray CsA treatment;blue LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SD. T-test. Two-way ANOVA test with Tukey’s multiple comparisons test, *p-value<0.05; ****p-value<0.00005; ns = non-significant.

E) Experimental design for EdU assay in regeneration. Larvae were treated with nifurpirinol for beta cell ablation from 3 to 4dpf then with CsA from 4 to 5dpf and then briefly treated with EdU before fixation and analysis.

F) Whole mount fluorescent immunohistochemistry (GFP and EdU) of the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 5dpf. 3D projection (stack) of one representative sample of non-ablated or ablated with or without CsA are shown. Scale 50µM.

G) Barplot representing the percentage of GFP+ ductal cells which incorporated EdU+ in pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 5dpf. The gray spheres represent non-ablated condition; the pink triangles the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SD. Two-way ANOVA test with Tukey’s multiple comparisons test, *p-value<0.05; ****p-value<0.00005; ns = non-significant.

H) Whole mount fluorescent immunohistochemistry (VenusPest and EdU) of the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(tp1:VenusPest) larvae at 5dpf. 3D projection (stack) of one representative sample of non-ablated or ablated with or without CsA are shown. Scale 50µM.

I) Barplot representing the percentage of GFP+ ductal cells which incorporated EdU+ in pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(tp1:VenusPest) larvae at 5dpf. The gray spheres represent non-ablated condition; the pink triangles the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± sd. Two-way ANOVA test with Tukey multiple comparisons test, *p-value<0.05; ****p-value<0.00005; ns means non-significant.

J) Barplot representing the number of VenusPest+ ductal cells which incorporated EdU+ in pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(tp1:VenusPest) larvae at 5dpf. The gray spheres represent non-ablated condition; the pink triangles the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SD. Two-way ANOVA test with Tukey’s multiple comparisons test, **p-value<0.005.

Incidentally, at 4dpf, while the proliferation is increased (Figure 5C), the number of ductal cells remained the same in all conditions (Figure 5B supplemental), suggesting that ductal cells have not yet left the cell cycle to differentiate. In comparison, after stronger Notch inhibition (15µM), the ductal cells are already depleted at 4dpf (Figure 5C supplemental), as they directly differentiate without entering the cell cycle [35]. In these conditions, CsA could therefore not enhance ductal progenitor proliferation and thus beta cell formation (Figure 4B). These results show that CaN and Notch pathways act together on the proliferation of the ductal progenitors to prevent their exhaustion.

CaN prevents exhaustion of Notch responsive progenitors during beta cell regeneration

Taken together, our results indicate that CaN plays a role in the proliferation of ductal progenitors in contexts that are permissive for beta cell differentiation. To demonstrate that CaN acts on ductal progenitor proliferation in a similar manner to Notch inhibition but during regeneration, we exposed Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae after ablation and CsA treatment (Figure 5E). CaN inhibition enhanced the proliferation of ductal cells in ablated larvae (Figure 5F-G). To next determine if CaN acts on the Notch-responsive progenitors in beta cell regeneration, we then used Tg(tp1:VenusPest) Notch reporter line. At larval stages the vast majority of ductal cells are Notch responsive and the reporter line marks all the progenitors within the ductal tree [34]. We treated Tg(ins:NTR-P2A-mCherry); Tg(tp1:VenusPest) as described above (Figure 5E). In ablated-larvae, CsA increased tp1+ ductal cell proliferation (Figure 5H-I). Moreover, CsA induced a reduction of tp1+ ductal cells in ablated larvae (Figure 5 H-J), suggesting an exhaustion of the Notch responsive progenitors, in accordance with premature beta cell differentiation we observed (Figure 2B-4D). Lastly, CsA does not affect tp1+ cells in non-ablated larvae showing that CaN inhibition did not directly affect Notch signaling (Figure 5J). Those results suggest that CaN fine tunes the balance between proliferation of the progenitors and their differentiation to prevent their exhaustion during beta cell regeneration.

CaN regulation is functionally relevant in adult zebrafish

To further expand upon our findings and investigate their relevance in a non-developmental context, we next ought to determine whether CaN function is maintained in older zebrafish. We used 2 months old Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) juveniles fish to perform beta cell ablation followed by CsA treatments. We analyzed the number of small islets (up to 5 cells) at 7 and 10dpt. At 7dpt, CsA increased the number of small islets in ablated juveniles (Figure 6A-B) showing that CaN inhibition enhances beta cell regeneration in juvenile zebrafish, as in larvae. It is noteworthy that, as in larvae, we highlighted an acceleration of beta cell regeneration. The increase number of small islets is indeed transient as it not observe anymore at 10dpt (Figure 6C).

CaN regulation is important in juveniles/adults and necessary for correct glycemia recovery

A) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) 2 months old zebrafish at 7dpt. 3D projection (stack) of non-ablated and ablated larvae treated with DMSO or CsA representative samples. The principal islet (PI) and the pancreatic tail are shown. Zoom from section of the pancreatic tail are shown, arrows point out mCherry+ beta cells in these zoom. Scale 200µM

B-C) Barplot representing the number of number of mCherry+ small secondary islets (=<5 cells) in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) 2 months old zebrafish at 7 (B) and 10dpt (C). The gray spheres represent non-ablated condition; the pink triangles represent the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SD. Two-way ANOVA with Tukey’s multiple comparison test, *p-value<0.05.

D-E) Barplot representing the glycemia (mg/dL) of Tg(ins:NTR-P2A-mCherry); months old adult zebrafish at 7 (D) and 10dpt (E). The pink triangles represent the ablated condition; the inverted green triangles ablated + CsA condition; the blue squares Tg(hsp70:CaNCA) after heat shocks; the orange lozenges Tg(UAS:CaNCA); Tg(cftr:gal4). The grey line represents the mean glycemia of controls (non-ablated) fish. Data are presented as mean values ± SD. One-way ANOVA with Tukey’s multiple comparison test, *p-value<0.05; **p-value<0.005.

Next, we determined the functional impact of CaN overexpression or inhibition by assessing the glycemia at 7, 10 and 14 after beta cell ablation. Overexpression of CaNCA using either Tg(hsp70:GFP-P2A-ppp3ccCA) or Tg(cftr:gal4); Tg(UAS:GFP-P2A-ppp3ccCA), led to an increased glycemia at both 7 and 10dpt (Figure 6D-E), before recovery at 14 dpt (Figure 6A supplemental). This indicates that the overexpression of CaN delayed glycemia recovery induced by beta cell regeneration. However, CaN inhibition did not seem to further improve the glycemia (Figure 6D-E) probably because the glycemia was already low at 7dpt (77 mg/dL on average) compared to non-ablated control (50 mg/dl on average) (Figure 6D-E). Altogether, these results show that in adult zebrafish also, CaN regulation is necessary to enable beta cell regeneration and for proper recovery of the glycemia after beta cell loss.

Discussion

The present study reveals the importance of the protein phosphatase CaN as a new player in beta cell regeneration in zebrafish. We demonstrate that CaN acts on ductal-associated progenitor cells by balancing proliferation and endocrine differentiation. In addition, we show the interplay between CaN and Notch signalling, a master regulator of beta cell regeneration.

Previous drug and genetic screening using zebrafish larvae enabled the identification of several regulators of beta cell regeneration from different pancreatic cellular sources. For example, adenosine has been shown to stimulate beta cell replication [28], igfbp1a [43] and TGFb suppression [30] promote alpha-to-beta cell transdifferentiation. As for cdk5 inhibition and folinic acid/Folr1, they promote beta cell regeneration from the pancreatic ducts [44], [45]. Here, to identify novel regulators of beta cell regeneration specifically from pancreatic ducts, we carried out a transcriptomic profiling of duct cells following beta cell ablation in the adult zebrafish. Transcriptomic analyses show that the regulated genes encompass most of the genes and pathways identified in previous studies (igfbp1, mTor, Notch, etc.), underlying the importance of those actors in beta cell regeneration. Our data reveal also that DNA replication is the most enriched signature attesting that duct cells undergo a potent proliferative response after the destruction of beta cells.

Besides these expected signatures, our transcriptomic data uncover the unanticipated upregulation of numerous genes implicated in DNA repair and cell cycle arrest. These signatures might indicate that highly proliferating ductal cells activate counteracting mechanisms. Among the genes regulated in those signatures, we focused on CaN and determined its role in beta cell regeneration. Pharmacological inhibition of CaN increases the proliferation of duct cells induced by beta cell ablation, resulting in an acceleration of beta cell regeneration in the ducts. Consistently, transgene-mediated CaN overactivation abolishes the regenerative response. Importantly, the inhibition of regeneration is observed when CaN is overexpressed either ubiquitously or selectively in cftr-expressing ductal cells indicating that the role of CaN in beta cell regeneration is intrinsic to the ducts. Based on functional assays in larvae, we not only confirm the activation of the proliferation of ductal cells soon after beta cell ablation, but also that the rate of progenitor proliferation is carefully controlled by CaN in order to achieve proper and timely regeneration of beta cells. Our data are consistent with earlier studies reporting a role of CaN in proliferation dynamics during fin regeneration. In the regenerating fin, low CaN activity is found in the proximal region of the blastema characterized by a high rate of proliferation and regeneration and its activity increases distally where lower proliferation is observed [36], [38], [39]. It was suggested that CaN control blastemal cell progeny divisions [36]. In human, the importance of CaN in proliferation is also highlighted in organ transplanted patients. When patients are treated with Cyclosporin A (i.e. the CaN inhibitor we used in this study) as immunosuppressive drug they indeed present an increased risk of skin cancer, notably due to keratinocyte senescence inhibition [46].

It has been shown that Notch inhibitory treatments switch progenitors from proliferative self-renewing to premature differentiation, leading to progenitor depletion [34], [35]. Our study reveals that this phenomenon is further exacerbated by CaN inhibition. Importantly, during normal larval development in absence of Notch inhibitory treatment, CaN does not affect basal ductal proliferation nor beta cell differentiation. Hence, Notch signalling has to be repressed to detect the effect of CaN on the progenitors, suggesting that CaN acts downstream of Notch pathway. In differentiating keratinocytes, CaN cooperates with Notch signalling to regulate p21/cdkn1a (which is upregulated in the ducts at 3 dpt), cell cycle withdrawal and differentiation [47]. These studies show that CaN acts in association with Notch signalling on progenitors proliferation and on their differentiation.

Based on our data, we build the model depicted in Figure 7A. Our study suggests that CaN acts in competent progenitor cells and that this competence is determined by Notch signalling. When Notch is repressed to a mild level, the progenitor enter into the cell cycle and acquire a pro-endocrinogenic competence [35]. CaN acts on these progenitors to tone down an excessive proliferation and avoid the exhaustion of these progenitors. CaN is therefore a guardian of the progenitor population. CaN inhibition both increases progenitor proliferation and induces their depletion, suggesting a switch to a symmetric division resulting in two daughter cells entering in endocrine differentiation.

Model of CaN action on ductal progenitors to regenerate beta cells

A) Under physiological conditions, the behavior of the ductal progenitors is determined by Notch signaling. Calcineurin is active in these progenitors and enable a proper control between proliferation and differentiation. When CaN in repressed, more ductal progenitors enter in the cell cycle (2dpt) and switch to a mode of proliferation leading to differentiation of the two daughter cells (4dpt), as more pro-endocrine cells are formed. The result is a exhaustion of the progenitors and a premature beta cell differentiation (10dpt).

More precisely, a previous study in mice uncover the existence of three division modes of the pancreatic progenitors during embryonic development, i.e. symmetric self-renewing resulting in two progenitors cells; asymmetric resulting in a progenitor and a endocrine cell and symmetric differentiative resulting in two differentiated cells [48]. The authors actually show that the type of division is defined by the timing of induction of endocrine program by NEUROG3 [48]. Interestingly, NEUROG3 seems to be the link between proliferation and differentiation, and its expression is regulated by Notch signalling [49]. In zebrafish, endocrine differentiation is not induced by Neurog3 but by Ascl1b and Neurod1 [50]. Concerning that subject, CaN inhibition accelerates the formation of neurod1+ cells. Hence CaN could possibly act via the determination of the type of division i.e. symmetric vs asymmetric. This model is supported by previous observation in others systems. In stem cells and neuronal and hematopoietic progenitors, premature differentiation results from a switch in the mode of cellular division, from symmetric amplifying division to asymmetric differentiating division [51], [52]. Notch determines the choice between both types of divisions [53], [54].

Calcineurin is known to be implicated in cellular senescence [46]. Usually thought as negative regulators of development and cellular growth, DNA repair [55] and cellular senescence [56] appear to be required in both developmental and regenerative processes. Our transcriptomic data suggest that these mechanisms are required for beta cell regeneration. Therefore, it would be interesting to determine the contribution of these cellular processes in the ductal progenitors and determine if CaN acts via cellular senescence in this case.

Overall, this study brings new insights on beta cell regeneration and highlights the ductal progenitor cell cycle as a cornerstone in the process. Some studies report an increase of proliferation of some ductal cell population in diabetic patients [10], [57], implying a regenerative response. However, these ductal cells cannot efficiently reform the beta cell mass, suggesting a dormant mechanism of regeneration. As such, the balance between proliferation and induction of endocrine differentiation could be a key to improve beta cell neogenesis. However, as CaN is also important for beta cell function, this approach would require to be transient to induce neogenesis. Therefore, it should be combined with methods to induce beta cell proliferation to ultimately reconstitute the beta cell mass. Overall, this study brings a better understanding on the regulation of the balance between ductal progenitors proliferation and endocrine differentiation. These results should provide new hints to help improve regenerative competences in mammals.

Material and Methods

Key resources table

Zebrafish husbandry and generation of the Tg(hsp70:eGFP-P2A-ppp3ccaCA)ulg068 and Tg(UAS:eGFP-P2A-ppp2ccaCA)ulg069 zebrafish lines

Tg BAC(nkx6.1:eGFP)ulg004 [58] ; Tg(ins:NTR-P2A-mCherry)ulg034 [59] ; Tg(cftr:gal4) and Tg(tp1:VenusPest) were used. Zebrafish were raised in standard conditions at 28°C. All experiments were carried out in compliance with the European Union and Belgian law and with the approval of the ULiège Ethical Committee for experiments with laboratory animals (approval number: 2075).

The hsp70:GFP-P2A-ppp3ccaCA transgene has been generated by cloning a PCR fragment containing the Gateway vector pCR8/GW/TOPO. Firstly, we amplified the full length of ppp3ccaCA with primers IM369/IM380 and amplified GFP-P2A with overlapping regions with IM371/O226. The overlapping PCR used the primers O226/IM380. Then to obtain a truncated ppp3ccaCA lacking the calmodulin biding and the autoinhibitory domain, resulting in a constantly active form of calcineurin, we amplified the last fragments with IM370/O226 and cloned into PCR8 vector. The promoter was assembled by LR recombination with pE5-hsp70 into pDestTol2p2A from the Tol2kit [60]. Tg (hsp70:GFP-P2A-ppp3ccaCA) fish have been generated using the Tol2 mediated transgenesis [61]. The Tg(UAS:GFP-P2A-ppp3ccacA) has been generated by ligation (KLD kit, NEB) of PCR fragments GFP-P2A-ppp3ccacA (IM459/IM460) with the UAS sequences (IM457/IM458) in plasmid from (Distel 2009) and then inserted into pDestTol2p2A from the Tol2kit. Final constructions has been injected with transposase into wild type (WT) AB embryos.

Beta-cell ablation

Adults fish for RNA-sequencing experiment were treated with freshly prepared metronidazole (MTZ) (Sigma M1547) at 10mM with 0.2% DMSO in fish water. Control treatments consisted of fish water containing 0.2% DMSO. Fish were treated for 18 hours in the dark. Nifurpirinol (NFP) (32439, Sigma-Aldrich) stock solution was dissolved at 2.5 mM in DMSO. b-cell ablation in Tg(nkx6.1:eGFP); Tg(ins:NTR-P2A-mCherry) larvae was induced by treatment with 2.5 µM NFP in E3. Control treatments consisted of E3 containing 0.2% DMSO. Larvae were treated for 18 hours in the dark.

Drug treatments

Cyclosporine A (Selleckchem, S2286), CHIR99021 and LY411575 (Sigma-Aldrich, SML0506) stock solution were dissolved at 10mM in DMSO. Larvae treatment were respectively performed at 0.1 µM and 5µM in E3. Control treatments consisted of E3 containing the same amount of DMSO than drug treatment. Larvae were treated for 18 hours in the dark.

5-ethynyl-2′-deoxyuridine (EdU) incorporation assay

Zebrafish larvae were incubated in 4 mM EdU dissolved in E3 water for 8 hours, they were then directly euthanised in tricaine and fixed in 4% PFA. EdU was detected according to the protocol of Click-iT™ EdU Cell Proliferation Kit for Imaging, Alexa Fluor™ 647 (ThermoFisher C10340) after whole mount immunodetection.

Heat Shock

Successive heat shocks of 30 minutes and 12 hours apart were performed at 39°C for larvae and 37°C for juveniles and adults zebrafish.

Whole mount immunodetection

Larvae were euthanized in tricaine and fixed in 4% PFA at 4 °C for IHC. The digestive tract of juveniles was dissected prior immunodetection and kept in methanol for at least 18 hours. After depigmentation with 3% H2O2/1% KOH, larvae were permeabilised 30 min in PBS/ Triton X-100 and incubated for two hours in blocking buffer (4% goat serum/1% BSA/PBS/0.1% Triton X-100). Primary and secondary antibodies were incubated at 4 °C overnight.

Primary antibodies: Living Colors Polyclonal anti-mCherry/dsRed (rabbit, 1:500, Clontech 632496), anti-GFP (chicken, 1:1000),

Secondary antibodies: Alexa Fluor-488, –568, –633 (goat, 1:750, Molecular Probes).

Flow cytometry and FACS

The whole pancreas from 3-4 fish of Tg(nkx6.1:eGFP); Tg(ins:NTR-P2A-mCherry) adult fish (6–10 months old, males and females) were dissected, collected and washed in HBSS without Ca2+/Mg2+. Live cell dissociation was performed in Tryple Select 1x solution (GIBCO) supplemented with 100 U/mL collagenase IV (Life Technologies 17104-019) and 40 µg/mL proteinase K (Invitrogen, 25530031) for 10 min at 28 °C.

The GFP+ cells were selected on FACS Aria III and sorted under purity mode and after exclusion of the doublets. The purity of the sorted cells was confirmed by epifluorescence microscopy (∼95 %). Cells (about 1000-5000/fish depending on the cell type) were immediately lysed with 0.5% Triton X-100 containing 2U/µl RNAse inhibitor and stored at –80 °C.

mRNA sequencing of FACSed cells and bioinformatic analyses

cDNAs were prepared from lysed cells according to SMART-Seq2.0 (Picelli et al., 2014) for low input RNA sequencing and libraries were prepared with Nextera® DNA Library kit (Illumina). Independent biological replicates of each cell type sequenced using Illumina NextSeq500 and obtained ∼20 million 100 bp paired-end reads. Reads were mapped and aligned to the zebrafish genome GRCz11 from Ensembl gene annotation version using STAR version 2.6.1 (Dobin et al., 2013). Gene expression levels were calculated with featureCounts (http://bioinf.wehi.edu.au/featureCounts/) and differential expression determined with DESeq2 (Love, Huber, & Anders, 2014). Expression values are given as normalized read counts. Poorly expressed genes with mean normalized expression counts <10 were excluded from the subsequent analyses. DESeq2 uses Wald test for significance with posterior adjustment of P values (Padj) using Benjamini and Hochberg multiple testing. The differentially expressed (DE) genes identified with a Padj cutoff of 0.05 were submitted for GO analysis using WebGestalt tool (Liao, Wang, Jaehnig, Shi, &Zhang, 2019).

Confocal microscopy and image analysis

Images were acquired using Leica SP5 confocal microscope. We used ImageJ to count the cells and Imaris to do the pictures.

Glycemia measurement

Glycemia measurement were performed as described in [59].

Data availability

The sequences that support the findings are of this study have been deposit in Geo with the accessions codes GSE212124, https://www.ncbi.nlm.nih.gov/geo/info/linking.html. The authors declare that all other data supporting the findings of this study are available within the paper and its supplementary information files.

Authors contribution

L.M. conducted experiments, acquired and analysed data, analysed transcriptomic data, wrote the manuscript. D.B. conducted RNA-seq experiment and revised the manuscript A.L. implemented bioinformatic programs and analysed transcriptomic data. C.R., C.G. and C.D. conducted experiments. L.F. revised the manuscript. B.P. provided resources and advised on research studies. M.L.V. provided resources, advised on research studies and revised the manuscript I.M. supervised the whole work, acquired funding, designed the research studies, verified data and assisted in writing the manuscript.

Supporting information

eLife Assessment

This work presents some valuable information regarding the molecular mechanisms controlling the regeneration of pancreatic beta cells following induced cell ablation. However, the study lacks the critical lineage tracing result to support the conclusion about the origin of the regenerated beta cells. The results of the pharmacological manipulation of CaN signaling are also incomplete. In particular, these manipulation are not cell-specific, making it difficult to interpret and thus a genetic approach is recommended.

Reviewer #1 (Public Review):

Induction of beta cell regeneration is a promising approach for the treatment of diabetes. In this study, Massoz et.al., identified calcineurin (CaN) as a new potential modulator of beta cell regeneration by using zebrafish as model. They also showed that calcineurin (CaN) works together with Notch signaling to promote the beta cell regeneration. Overall, the paper is well organized, and technically sound. However, some evidences seem weak to get the conclusion.

Reviewer #2 (Public Review):

This work started with transcriptomic profiling of ductal cells to identify the upregulation of calcineurin in the zebrafish after beta-cell ablation. By suppressing calcineurin with its chemical inhibitor cyclosporin A and expressing a constitutively active form of calcineurin ubiquitously or specifically in ductal cells, the authors found that inhibited calcineurin activity promoted beta-cell regeneration transiently while ectopic calcineurin activity hindered beta-cell regeneration in the pancreatic tail. They also showed similar effects in the basal state but only when it was within a particular permissive window of Notch activity. To further investigate the roles of calcineurin in the ductal cells, the authors demonstrated that calcineurin inhibition additionally induced the proliferation of the ductal cells in the regenerative context or under a limited level of Notch activity. Interestingly, the enhanced proliferation was followed by a depletion of ductal cells, suggesting that calcineurin inhibition would exhaust the ductal cells. Based on the data, the authors proposed a very attractive and intriguing model of the role of calcineurin in maintaining the balance of the progenitor proliferation and the endocrine differentiation. However, the conclusions of this paper are only partially supported by the data as some evidence of the lineage between ductal cells and beta cells remains suggestive.

Author response:

The following is the authors’ response to the previous reviews.

Thank you for all your recommendations to improve the manuscript. We took them into account and tried to integrate them as much as possible in the paper. I understand that the main issue is the lack of genetic lineage tracing. Unfortunately, I am no longer in a position to perform experiments and as a consequence, we cannot bring these data. However, we previously performed several experiments that attest the ductal origin of the beta cells. As a reminder, we used experiment setting where beta cell regeneration occur from the ducts in the pancreatic tail; we used a genetic approach to over-express CaN specifically in the ducts at the level of the pancreas ; and we investigate the function of CaN under Notch repression, known to trigger beta cell formation from the ducts. Altogether, our data underline the contribution of the ductal cells. In addition, as recommended by the editors, we showed that while the proportion of ductal cells EdU+ increase Figure 5 C-D, the number of ductal cells remain constant Figure 5A supplemental. We integrate a paragraph in the discussion to remind all these points in the manuscript.

We thank you greatly for your time and consideration for this work.

Version 3

Abstract

Abstract

Stimulation of pancreatic beta cell regeneration could be a therapeutic lead to treat diabetes. Unlike humans, the zebrafish can efficiently regenerate beta cells, notably from ductal pancreatic progenitors. To gain insight into the molecular pathways involved in this process, we established the transcriptomic profile of the ductal cells after beta cell ablation in the adult zebrafish. These data highlighted the protein phosphatase calcineurin as a new potential modulator of beta cell regeneration. We showed that calcineurin overexpression abolished the regenerative response, leading to glycemia dysregulation. On the opposite, calcineurin inhibition increased ductal cell proliferation and subsequent beta cell regeneration. Interestingly, the enhanced proliferation of the progenitors was paradoxically coupled with their exhaustion. This suggests that the proliferating progenitors are next entering in differentiation. Calcineurin appears as a guardian which prevents an excessive progenitor proliferation to preserve the pool of progenitors. Altogether, our findings reveal calcineurin as a key player in the balance between proliferation and differentiation to enable a proper beta cell regeneration.

Introduction

Blood glucose homeostasis is tightly controlled by pancreatic endocrine cells. Insulin-producing beta cells work in close association with alpha cells, which secrete glucagon, and delta cells secreting somatostatin, to maintain normal glycemia. The insulin plays a critical role in this process as it is the only hormone able to lower the glycemia. Beta cell loss is a hallmark of type 1 and of late stages of type 2 diabetes, leading to chronic hyperglycemia. Beta cell destruction is largely irreversible in human and in mammals, making the disease incurable nowadays. Nevertheless, studies in diabetic mice models uncovered promising evidences of slight recovery of beta cells via regeneration. For example, new beta cells can arise from the replication of remaining beta cells [1]. Different models of pancreas injuries revealed the plasticity of mammalian pancreatic cells. Actually, differentiated endocrine cells such as alpha cells[2] and delta cells[3] can reprogram and convert into insulin-producing cells, a process that is age dependent. Another possible source of beta cells could be de novo formation from pancreatic progenitors residing in the ductal compartment of the adult pancreas [4]. Lineage tracing experiments could not clearly highlight new beta cells arising from the ductal cells in the adult [4]–[6]. However, expression of the pro-endocrine marker Neurog3 was detected in the ducts in different mouse models of pancreas regeneration [4], [7], [8]. In addition, a rare population of ductal cells expressing Neurog3 has been reported to contribute to beta cell neogenesis during diabetes [9]. A recent study of single cell RNA sequencing reveal a sub-population of human ductal cells which are able to give rise to all the pancreatic cell types, including beta cells, following implantation in mice [10]. Altogether, these findings suggest that mammalian pancreatic ducts possess the intrinsic capacity to (re)generate beta cells even though this process is poorly efficient and slow, especially in adults. In contrast to mammals, zebrafish possess remarkable capacity of regeneration, independently of its age [11]–[16]. This model can therefore be exploited to identify and characterize regenerative mechanisms and ultimately induce regeneration in mammals. Based on regenerative mechanisms identified in zebrafish, several studies succeed to improve regeneration in mammals, underlying the possibility of translation from zebrafish to mammals ([17]–[21] and [22], for review).

In zebrafish, the nitroreductase (NTR)/nitroaromatic prodrug system is widely used. In this technique, the combination of cell type-specific NTR expression and nitroaromatic prodrug exposure allow for controlled and targeted cell ablation [23], [24]. When NTR is expressed under the control of the insulin (ins) promoter (Tg(ins:NTR-mCherry) [25]), beta cell destruction is complete within 3 days following the treatment in adult fish, which correlate with a peak of hyperglycemia [26], [27]. The regeneration of beta cells upon ablation is spontaneous and fast and the glycemia is normalized within two weeks [13], [29]. Similar to mice, new beta cells can arise through proliferation of surviving beta cells [28] as well as through the contribution of alpha cells [29], [30] and delta cells[31], [32], underscoring the overall conservation of these processes from zebrafish to mammals. Nevertheless, unlike mammals, the presence of pancreatic progenitors in the ducts is well established in both larval [33] and adult zebrafish [26], [27]. Lineage tracing experiments pointed out that ductal cells and centroacinar cells (CACs) give rise to new beta cells [26], [33]. In adults, duct-derived beta cells start to be detected between 7 and 10 days following beta cell ablation [26], [27].

Notch signaling is a key player regulating the differentiation of duct-associated progenitors into endocrine cells. Larval duct cells as well as adult CACs display strong Notch activity [26], [34]. This signaling pathway has a central role in beta cell genesis during both development [35] and regeneration [33] by repressing endocrine differentiation. In zebrafish, different levels of Notch activity determine the behavior of the pancreatic progenitors. While a high level of Notch activity maintains cells in quiescence, a moderate level induces the entry in the cell cycle and proliferation whereas a low level drives endocrine differentiation of the progenitor cells [35]. A steep decrease of Notch activity pushes the progenitors to differentiate prematurely, bypassing the amplification step and leading to their depletion [35]. Repression of the Notch signaling by mTor, activated by glucose and nutrients, hence promotes beta cell formation and regeneration from ductal progenitors [33]. While Notch and mTor signaling are crucial for this process, there is still a need to establish a global view of the molecular mechanisms regulating beta cell regeneration.

To identify early events regulating ductal-derived beta cell regeneration, we determined the transcriptomic signature of ductal cells from adult zebrafish following beta cell destruction. Our data highlighted an upregulation of the Calcineurin (CaN) pathway. To elucidate CaN function in beta cell regeneration, we both repressed and activated CaN pathway. We showed that CaN regulates beta cell neogenesis in the ducts during regeneration, by acting on progenitor proliferation. Together, our findings underline that CaN fine tunes the balance between progenitor proliferation and beta cell differentiation to guarantee proper regeneration.

Results

Transcriptomic profiling of ductal cells after beta cell destruction highlights regulation of calcineurin pathway

To gain a better understanding of the molecular mechanisms underlying the regeneration of beta cells from the ducts, we determined the transcriptional landscape of ductal cells by RNA-sequencing after beta cell ablation in the adult zebrafish. To selectively ablate the beta cells, we used the Tg(ins:NTR-mCherry) transgenic fish. The ductal cells were labelled thanks to the Tg(nkx6.1:GFP) reporter line [27] in which GFP marks the ductal tree and associated multipotent pancreatic progenitors [27]. More precisely, 3 to 4 fishes were treated with the pro-drug MTZ at 10mM overnight, to induce ablation of beta cells or with DMSO for the non-ablated controls. Ablation was confirmed by blood glucose measurement before collection of the pancreas. To capture the early events triggered by the destruction of beta cells, we generated the transcriptome of the ducts 3 days post ablation treatment (dpt), i.e. before beta cell neogenesis (Figure 1A supplemental). Differential gene expression analysis revealed that 1866 genes are up-regulated and 1515 genes down-regulated in the ductal cell of fish treated with MTZ compared to control (pAdj < 0.05). According to Gene Ontology analysis, the most enriched pathways among the up-regulated genes were DNA replication and cell cycle (Figure 1A). This further corroborates our previous findings and those of others, regarding the activation of duct-associated progenitors’ proliferation in response to beta cell ablation [26], [27]. As expected, the Notch pathway was enriched in the down-regulated genes [33], [35] (Figure 1B). On a contrasting note, in a context marked by a robust proliferative response, we found it intriguing to observe an enrichment in the cellular senescence signature (Figure 1A). Subsequently, we investigated deeper into the genes associated with this specific signature. We found among them several components of the calcineurin signaling pathway such as nfatc3b, ppp3ccb (the catalytic subunit of Calcineurin), itpr2 and calm3b (Figure 1C). In addition to these genes related to the cellular senescence signature, our transcriptomic studies revealed the modulation of other genes from the CaN canonical pathway (Figure 1D), underlying its potential role in beta cell regeneration. Calcineurin (CaN) is a highly conserved calcium/calmodulin dependent Ser/Thr phosphatase, involved in numerous biological process including fin regeneration and beta cell function [36]–[39]. This prompted us to investigate the role of CaN in beta cell regeneration.

Transcriptomic profiling of ductal cells during beta cell regeneration and validation in larvae.

A-B) Enrichment ratio of selected non-redundant signatures of KEGG pathways overrepresented in ductal cells after beta cells ablation (UP – A and DOWN-B) compared to ductal cells without beta cells ablation. Gene Ontology (GO) terms were identified using ORA analysis by WebGestalt (Liao et al., 2019) using the list of DE genes provided by DESeq. The light color for Notch pathway means p-value = 0.11.

C) List of genes associated with the signature of cellular senescence from A-B. Genes related to CaN pathway are in bold.

D) Calcineurin canonical pathway with up-regulated genes in green and down-regulated genes in red in transcriptomic data from A-B.

E) Experimental design for regeneration test in larvae. Briefly, after nifurpirinol treatment from 3 to 4dpf, larvae were fixed and analyzed at 4-7-10 and 14 days post treatment (dpt).

F) Graph representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) at 0-4-7-10 and 14 dpt. The gray spheres represent non-ablated conditions and the pink triangles the ablated condition. Data are presented as mean values ± SEM. One-way ANOVA test with Tukey’s multiple comparison test, ****p-value<0.0005.

G) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 14dpt. 3D projection (stack) of one non-ablated and one ablated representative samples. The principal islet (PI) and the pancreatic tail are shown. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 100µM.

H) Experimental design for EdU assay in larvae. After NFP treatment for 3 to 4dpf, larvae were exposed to EdU at 2dpt before fixation for analysis.

I) Whole mount fluorescent immunohistochemistry (GFP and EdU) of the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 2dpt. 3D projection (stack) of one non-ablated and one ablated representative samples. Arrows point out GFP+ duct cells EdU+ in the pancreatic tail. Scale 50µM.

J) Barplot representing the percentage of GFP+ ductal cells which incorporated EdU+ in non-ablated and ablated conditions. Data are presented as mean values ± SD. T-test. **p-value<0.005.

Previous RNAseq data performed in our laboratory indicate that CaN (ppp3cca/b) and NFATc3 (nfatc3a/b) are mainly express in endocrine cells [40] (Figure 1B supplemental), which is in accordance with the role of CaN/NFAT signalling in beta cells [41]. CaN genes (ppp3cca/b) as well as nfatc3b, are express at lower levels in the ducts at basal state but their expression is induced in response to beta cell destruction (Figure 1C supplemental).

Calcineurin activity regulates the ductal regenerative response

To evaluate the role of CaN in beta cell regeneration, more specifically derived from ductal progenitors, we have chosen to use young larvae, where regenerated beta cells in the pancreatic tail arise exclusively from the ducts [33]. In response to beta cells ablation, the intrapancreatic ducts undergo a ductal regenerative response whereby differentiation toward the endocrine fate is increased [33]. We first determined the rate of beta cell neogenesis from the ducts in response to a single acute ablation of beta cells as we performed in adults. We treated Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae with nifurpirinol (NFP) from 3 to 4 dpf and mCherry+ beta cells were quantified in the GFP+ ducts in the pancreatic tail at several time points: 4, 7, 10 and 14 dpt (Figure 1E). Duct-associated beta cells started to be detected in non-ablated larvae between 7 and 10 dpt (Figure 1F) and the number of beta cells slowly increased until 14 dpt (Figure 1F, 1G). In ablated larvae, the increase became more pronounced from 10 dpt onwards (Figure 1F, 1G), indicating faster endocrine differentiation. This experiment establishes that the ductal regenerative response is detectable between 10 and 14 days after the beta cell ablation, performed at 3 dpf. We next wanted to determine if ductal cell proliferation is activated in response to beta cell destruction in larvae as in adult fish. We exposed Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae to EdU the second day following ablation (Figure 1H). In the ablated larvae, the proportion of GFP+ ductal cells EdU+ (in S-phase) was higher compared to non-ablated larvae (Figure 1I, 1J). This result shows that acute beta cell ablation in larvae rapidly activates ductal cell proliferation, as previously reported in adult zebrafish.

As our transcriptomic data from adult zebrafish revealed modulation of the CaN pathway at 3 dpt (Figure 1D), we treated Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae from 1 to 3 dpt after beta cell ablation, with a CaN inhibitor, the Cyclosporin A (CsA) [39] (Figure 2A). The number of newly formed beta cells in the tail was monitored from 4 to 14 dpt. CsA enhanced beta cell formation at 10 dpt (Figure 2B-D). However, this effect appears to be transient since no discernible difference was observed between the control and CsA-treated larvae in regeneration at the latest time point (14dpt), suggesting an acceleration of the regenerative response (Figure 2B, Figure 2A supplemental). Interestingly, CsA did not affect beta cell differentiation in non-ablated larvae, indicating that CsA only acts in a regenerative context (Figure 2B-D). Of note, CsA increased as well the number of regenerated beta cells in the principal islet (Figure 2B, 2C supplemental). We and others, previously showed that the other major source of beta cell regeneration, besides the ducts, are the bi-hormonal sst1.1+/ins+ cells [31], [32], [42]. However, CsA did not affect bi-hormonal cell formation (Figure 2D supplemental), suggesting that the additional cells in the principal islet could also originate from the ducts.

Calcineurin inhibition with CsA increases the ductal regenerative response.

A) Experimental design for regeneration test in larvae with CsA treatment. Briefly, after nifurpirinol treatment from 3 to 4dpf, larvae were treated with CsA from 1 to 3dpt and fixed and analyzed at 4-7-10 and 14 days post treatment (dpt).

B) Graph representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) at 0-4-7-10 and 14 dpt. The gray spheres represent non-ablated condition; the pink triangles represent the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SEM. Two-way ANOVA test with Sidak’s multiple comparisons test, *p-value<0.05.

C) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 10dpt. 3D projection (stack) of non-ablated and ablated larvae treated with DMSO or CsA representative samples. The principal islet (PI) and the pancreatic tail are shown. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 100µM.

D) Barplot representing the number of number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 10dpt. The gray spheres represent non-ablated condition; the pink triangles represent the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SD. Two-way ANOVA with Tukey’s multiple comparison test, *p-value<0.05.

E) Graph representing the mean number of GFP+ neurod1+ cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(neurod1:GFP) at 0-4-7 and 10 dpt. The gray spheres represent non-ablated condition; the pink triangles represent the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SEM. Two-way ANOVA test with Sidak’s multiple comparisons test, **p-value<0.005; *** p-value<0.0005; **** p-value<0.00005.

F) Graph representing the mean number of GFP+ neurod1 EdU+ cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(neurod1:GFP) at 0-4-7 and 10 dpt. The gray spheres represent non-ablated condition; the pink triangles represent the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SEM.

We next tested the effect of CsA on endocrine progenitors in a regenerative context. We induced regeneration in Tg(neurod1:GFP) larvae where the GFP is expressed in both endocrine progenitors and mature endocrine cells. We first assessed generation of GFP cells at different time points (Figure 2E supplemental) and showed that CsA induced an increase of neurod1+ cells from 4 dpt. The increase was still detectable at least until 10dpt (Figure 2E, Figure 2F supplemental). To determine if these additional cells result from their own proliferation, we performed a pulse of EdU just before analysis (Figure 2E supplemental). We observed that CsA did not affect the neurod1+ cell proliferation rate, which is very low at these stages (Figure 2F). As a consequence, the effect of CsA cannot be explained by endocrine cell proliferation but rather by neogenesis from progenitors. As CsA affects pro-endocrine cells formation, we next wondered if the increased cell formation induced by CsA is specific to beta cells. Treatment with CsA was performed as previously and delta1.1 and alpha cells were detected by immunofluorescence. Interestingly, CsA did not affect alpha nor delta1.1 cells neogenesis in response to beta cell ablation (Figure 2G-H supplemental). Overall, these experiments showed that CsA affects specifically the beta cells and their endocrine progenitors.

CaN over-activation abolishes the regenerative response

We then wondered if an opposite regulation of CaN i.e. its activation impacts as well the regenerative response. To that end, we generated a transgenic line Tg(hsp70:GFP-P2A-ppp3ccCA) that allows ubiquitous expression of a constitutively active form of CaNCA (ppp3ccCA) upon heat shocks. Beta cell ablation was triggered in Tg(hsp70:GFP-P2A-ppp3ccCA); Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae from 3 to 4 dpf and CaNCA expression was induced by four successive heat-shocks from 1 to 3 dpt (Figure 3A). The overexpression of CaNCA after ablation impaired the regenerative response at 14 dpt (Figure 3B-C). Similar results were obtained with Tg(UAS:GFP-P2A-ppp3ccCA); Tg(cftr:gal4)[45]; Tg(ins:NTR-P2A-mCherry) larvae in which CaNCA is continuously and specifically overexpressed in the ducts within the pancreas (Figure 3D-E). Importantly, the structure of the ducts was similar in CaNCA overexpressing larvae compared to Tg(nkx6.1:GFP) controls (Figure 3D), suggesting that the suppression of the regenerative response in CaNCA-overexpressing larvae was not due to morphogenetic defects during ductal growth. This result shows that CaN acts directly in the ducts to decrease beta cell regeneration while it is not necessary for normal beta cell differentiation.

Transgenic mediated overexpression of calcineurin abolish the ductal regenerative response.

A) Experimental design for regeneration test in larvae with heat-shocks. Briefly, after nifurpirinol treatment from 3 to 4dpf, four heat shock were performed from 1 to 3dpt and larvae were fixed and analyzed at 14dpt.

B) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(hsp70:CaNCA); Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 14dpt. 3D projection (stack) of one non-ablated and one ablated with or without heat-shock representative samples. The principal islet (PI) and the pancreatic tail are showed. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 100µM.

C) Barplot representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(hsp70:CaNCA); Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 14dpt. The gray spheres represent non-ablated condition; the pink triangles represent the ablated condition; the black squares heat-shock condition and inverted green triangles ablated + heat-shock condition. Data are presented as mean values ± SD. Two-way ANOVA with Tukey’s multiple comparisons test, *p-value<0.05, *** p-value<0.0005, ns = non-significant.

D) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of larvae at 14dpt. 3D projection (stack) of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) one non-ablated and one ablated representative control samples and Tg(UAS:CaNCA); Tg(ins:NTR-P2A-mCherry); Tg(cftr:gal4) one non-ablated and one ablated representative samples. The principal islet (PI) and the pancreatic tail are showed. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 100µM.

E) Barplot representing the mean number of mCherry+ beta cells in the pancreatic tail of larvae at 14dpt. The gray spheres represent non-ablated Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) condition; the pink triangles represent the ablated Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) condition; the black squares non-ablated Tg(UAS:CaNCA); Tg(ins:NTR-P2A-mCherry); Tg(cftr:gal4) condition and inverted green triangles ablated Tg(UAS:CaNCA); Tg(ins:NTR-P2A-mCherry); Tg(cftr:gal4) condition. Data are presented as mean values ± SD. Two-way ANOVA with Tukey’s multiple comparison test, *** means p-value<0.0005, **** p-value<0.00005, ns = non-significant.

CaN regulates beta cell differentiation induced by Notch inhibition in absence of regeneration

Our transcriptomic data showed that the Notch pathway is downregulated in ductal cells during beta cell regeneration (Figure 1). As the level of Notch activity determines the behavior of ductal cells [33] from quiescence to proliferation and subsequently to beta cell differentiation, we tested whether calcineurin acts together with the Notch pathway on a common pool of ductal progenitors. To inhibit the Notch pathway, we treated larvae with several concentrations of the gamma-secretase inhibitor LY411575 from 3 to 4dpf in absence of regeneration. The activity of CaN was inhibited by CsA during the same timeframe (Figure 4A). As previously, we used reporter lines for beta and ductal cells Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) and the number of secondary beta cells was analyzed at 6dpf (Figure 4A). As expected, the number of beta cells progressively rose as the concentration of the Notch inhibitor increased (Figure 4B). Combined treatment with CsA fostered the differentiation of beta cells between 1 and 10 µM LY411575 but did not result in further increase at 15 µM LY411575 (Figure 4B), suggesting that CaN is important within a permissive window of Notch activity. Since we overserved the highest synergistic effect at 5µM of LY411575, we used this concentration for the following experiments (Figure 4C-D). It is worth noting that combined treatment of LY411575 (5uM) and another CaN inhibitor, FK506, resulted in the same synergistic increase of beta cell differentiation (Figure 4A-B supplemental), confirming that the effect is well due to CaN inhibition. The combined effect of Notch and CaN inhibition is transient as it is not observed at 7dpf anymore (Figure 4E). Therefore, as observed in regenerative conditions (Figure 2B), CaN inhibition accelerates beta cell neogenesis induced by Notch repression.

CaN repression potentializes the effect of Notch inhibition on beta cell formation.

A) Experimental design for Notch inhibition test in non-ablated condition. Larvae were treated concomitantly with LY411575 (Notch inhibitor) and CsA from 3 to 4dpf and were fixed and analyzed at 6dpf.

B) Graph representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 6 dpf depending the concentration of LY411575. The blue dots represent LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SEM. Two-way ANOVA test with Sidak’s multiple comparison test, *p-value<0.05, **p-value<0.05, ns = non-significant.

C) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 6dpf. 3D projection (stack) of one control (without any treatment); one CsA-treated; one LY411757-treated and one with both CsA and LY411575 treated larvae. The principal islet (PI) and the pancreatic tail are showed. Arrows point out mCherry+ beta cells in the pancreatic tail. Scale 50µM.

D) Barplot representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 6dpf. The black dots represent the control; gray CsA treatment;blue LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SD. Two-way ANOVA with Tukey multiple comparison test, ***p-value<0.0005, ****p-value<0.00005.

E) Graph representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 5-6-7 dpf. The black dots represent the control; gray CsA treatment;blue LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SEM. Two-way ANOVA test with Sidak’s multiple comparison test, *p-value<0.05.

To determine to which extent CaN pathway can modulate Notch induced beta cell neogenesis, we activated CaNCA overexpression in Tg(hsp70:GFP-P2A-ppp3ccCA); Tg(ins:NTR-P2A-mCherry) larvae by an heat-shock at 3 dpf and treated them with LY411575 (Figure 4C supplemental). CaNCA overexpression resulted in a lowered beta cell formation induced by Notch inhibition (Figure 4D supplemental), revealing that CaN activation counterbalanced the effects of Notch inhibition. Using our previous settings of Notch and CaN inhibition, we next wondered if the canonical pathway downstream of CaN was involved in the enhancement of beta cell differentiation. To activate NFAT, a well-known target of CaN, we used CHIR99021 allowing a stabilization of the active form of NFAT (Figure 1D). We found that CHIR99021 rescued the effect of CsA (Figure 4E supplemental), suggesting that CaN inhibition increases beta cell neogenesis at least partially by the regulation of NFAT. Overall, these results reveal that CaN impacts beta cell formation in pro-endocrinogenic context, such as induced by a low level of Notch activity. Moreover, it suggests that both CaN and Notch pathways act on a common pool of ductal progenitors to govern beta cell neogenesis.

CaN controls the proliferation of duct-associated progenitors induced by Notch inhibition

Given that we observed the most significant increase in beta cell formation with CsA when Notch activity was mildly repressed (Figure 4B-D), and since mild Notch activity has been shown to promote progenitor amplification [35], it suggests that CaN acts at this level. To explore this possibility, we exposed briefly Tg(nkx6.1:GFP) larvae to EdU after mild Notch (LY411575 5µM) and CaN inhibition (Figure 5A) and analyzed at 4dpf and 6dpf. As expected, Notch inhibition increased the proportion of proliferating EdU+ GFP+ ductal cells at 4dpf (Figure 5B-C) while the number of GFP+ ductal cells remained constant (Figure 5A supplemental). Furthermore, the amount of ductal progenitors decreased 2 days later (at 6dpf) (Figure 5D), which is concomitant with the increase of beta cell differentiation (Figure 4C-D). Interestingly, the combined inhibition with CsA further increased these proportions, while CaN inhibition alone had no effect (Figure 5B-D). Of note, we detected beta cells EdU+ at 6dpf (Figure 5B supplemental), underlying that these cells originate from the proliferation of ductal progenitors.

CaN repression increases the proportion of ductal proliferating cells.

A) Experimental design for EdU assay in Notch test. Larvae were treated concomitantly with LY411575 (Notch inhibitor) and CsA from 3 to 4dpf and then briefly treated with EdU before fixation and analysis at 4 dpf or at 6dpf.

B) Whole mount fluorescent immunohistochemistry (GFP and EdU) of the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 4dpf. 3D projection (stack) of one control (without any treatment), one with CsA only, one with LY411575 only and one with both CsA and LY411757 representative samples. Scale 50µM.

C) Barplot representing the percentage of GFP+ ductal cells which incorporated EdU+ in pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae for the Notch test. The black dots represent the control; gray CsA treatment;blue LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SD. T-test. Two-way ANOVA test with Tukey’s multiple comparisons test, *p-value<0.05; ****p-value<0.00005; ns = non-significant.

D) Barplot representing the number of GFP+ ductal cells which in pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 6dpf for the Notch test. The black dots represent the control; gray CsA treatment;blue LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SD. T-test. Two-way ANOVA test with Tukey’s multiple comparisons test, *p-value<0.05; ****p-value<0.00005; ns = non-significant.

E) Experimental design for EdU assay in regeneration. Larvae were treated with nifurpirinol for beta cell ablation from 3 to 4dpf then with CsA from 4 to 5dpf and then briefly treated with EdU before fixation and analysis.

F) Whole mount fluorescent immunohistochemistry (GFP and EdU) of the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 5dpf. 3D projection (stack) of one representative sample of non-ablated or ablated with or without CsA are shown. Scale 50µM.

G) Barplot representing the percentage of GFP+ ductal cells which incorporated EdU+ in pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 5dpf. The gray spheres represent non-ablated condition; the pink triangles the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SD. Two-way ANOVA test with Tukey’s multiple comparisons test, *p-value<0.05; ****p-value<0.00005; ns = non-significant.

H) Whole mount fluorescent immunohistochemistry (VenusPest and EdU) of the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(tp1:VenusPest) larvae at 5dpf. 3D projection (stack) of one representative sample of non-ablated or ablated with or without CsA are shown. Scale 50µM.

I) Barplot representing the percentage of GFP+ ductal cells which incorporated EdU+ in pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(tp1:VenusPest) larvae at 5dpf. The gray spheres represent non-ablated condition; the pink triangles the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± sd. Two-way ANOVA test with Tukey multiple comparisons test, *p-value<0.05; ****p-value<0.00005; ns means non-significant.

J) Barplot representing the number of VenusPest+ ductal cells which incorporated EdU+ in pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(tp1:VenusPest) larvae at 5dpf. The gray spheres represent non-ablated condition; the pink triangles the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SD. Two-way ANOVA test with Tukey’s multiple comparisons test, **p-value<0.005.

Incidentally, at 4dpf, while the proliferation is increased (Figure 5C), the number of ductal cells remained the same in all conditions (Figure 5A supplemental), suggesting that ductal cells have not yet left the cell cycle to differentiate. In comparison, after stronger Notch inhibition (15µM), the ductal cells are already depleted at 4dpf (Figure 5C supplemental), as they directly differentiate without entering the cell cycle [35]. In these conditions, CsA could therefore not enhance ductal progenitor proliferation and thus beta cell formation (Figure 4B). These results show that CaN and Notch pathways act together on the proliferation of the ductal progenitors to prevent their exhaustion.

CaN prevents exhaustion of Notch responsive progenitors during beta cell regeneration

Taken together, our results indicate that CaN plays a role in the proliferation of ductal progenitors in contexts that are permissive for beta cell differentiation. To demonstrate that CaN acts on ductal progenitor proliferation in a similar manner to Notch inhibition but during regeneration, we exposed Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae after ablation and CsA treatment (Figure 5E). CaN inhibition enhanced the proliferation of ductal cells in ablated larvae (Figure 5F-G). To next determine if CaN acts on the Notch-responsive progenitors in beta cell regeneration, we then used Tg(tp1:VenusPest) Notch reporter line. At larval stages the vast majority of ductal cells are Notch responsive and the reporter line marks all the progenitors within the ductal tree [34]. We treated Tg(ins:NTR-P2A-mCherry); Tg(tp1:VenusPest) as described above (Figure 5E). In ablated-larvae, CsA increased tp1+ ductal cell proliferation (Figure 5H-I). Moreover, CsA induced a reduction of tp1+ ductal cells in ablated larvae (Figure 5 H-J), suggesting an exhaustion of the Notch responsive progenitors, in accordance with premature beta cell differentiation we observed (Figure 2B-4D). Lastly, CsA does not affect tp1+ cells in non-ablated larvae showing that CaN inhibition did not directly affect Notch signaling (Figure 5J). Those results suggest that CaN fine tunes the balance between proliferation of the progenitors and their differentiation to prevent their exhaustion during beta cell regeneration.

CaN regulation is functionally relevant in adult zebrafish

To further expand upon our findings and investigate their relevance in a non-developmental context, we next ought to determine whether CaN function is maintained in older zebrafish. We used 2 months old Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) juveniles fish to perform beta cell ablation followed by CsA treatments. We analyzed the number of small islets (up to 5 cells) at 7 and 10dpt. At 7dpt, CsA increased the number of small islets in ablated juveniles (Figure 6A-B) showing that CaN inhibition enhances beta cell regeneration in juvenile zebrafish, as in larvae. It is noteworthy that, as in larvae, we highlighted an acceleration of beta cell regeneration. The increase number of small islets is indeed transient as it not observe anymore at 10dpt (Figure 6C).

CaN regulation is important in juveniles/adults and necessary for correct glycemia recovery.

A) Whole mount fluorescent immunohistochemistry (GFP and mCherry) of the pancreas of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) 2 months old zebrafish at 7dpt. 3D projection (stack) of non-ablated and ablated larvae treated with DMSO or CsA representative samples. The principal islet (PI) and the pancreatic tail are shown. Zoom from section of the pancreatic tail are shown, arrows point out mCherry+ beta cells in these zoom. Scale 200µM

B-C) Barplot representing the number of number of mCherry+ small secondary islets (=<5 cells) in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) 2 months old zebrafish at 7 (B) and 10dpt (C). The gray spheres represent non-ablated condition; the pink triangles represent the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SD. Two-way ANOVA with Tukey’s multiple comparison test, *p-value<0.05.

D-E) Barplot representing the glycemia (mg/dL) of Tg(ins:NTR-P2A-mCherry); months old adult zebrafish at 7 (D) and 10dpt (E). The pink triangles represent the ablated condition; the inverted green triangles ablated + CsA condition; the blue squares Tg(hsp70:CaNCA) after heat shocks; the orange lozenges Tg(UAS:CaNCA); Tg(cftr:gal4). The grey line represents the mean glycemia of controls (non-ablated) fish. Data are presented as mean values ± SD. One-way ANOVA with Tukey’s multiple comparison test, *p-value<0.05; **p-value<0.005.

Next, we determined the functional impact of CaN overexpression or inhibition by assessing the glycemia at 7, 10 and 14 after beta cell ablation. Overexpression of CaNCA using either Tg(hsp70:GFP-P2A-ppp3ccCA) or Tg(cftr:gal4); Tg(UAS:GFP-P2A-ppp3ccCA), led to an increased glycemia at both 7 and 10dpt (Figure 6D-E), before recovery at 14 dpt (Figure 6A supplemental). This indicates that the overexpression of CaN delayed glycemia recovery induced by beta cell regeneration. However, CaN inhibition did not seem to further improve the glycemia (Figure 6D-E) probably because the glycemia was already low at 7dpt (77 mg/dL on average) compared to non-ablated control (50 mg/dl on average) (Figure 6D-E). Altogether, these results show that in adult zebrafish also, CaN regulation is necessary to enable beta cell regeneration and for proper recovery of the glycemia after beta cell loss.

Discussion

Previous drug and genetic screening using zebrafish larvae enabled the identification of several regulators of beta cell regeneration from different pancreatic cellular sources. For example, adenosine has been shown to stimulate beta cell replication [28], igfbp1a [43] and TGFb suppression [30] promote alpha-to-beta cell transdifferentiation. As for cdk5 inhibition and folinic acid/Folr1, they promote beta cell regeneration from the pancreatic ducts [44], [45]. Here, to identify novel regulators of beta cell regeneration specifically from pancreatic ducts, we carried out a transcriptomic profiling of duct cells following beta cell ablation in the adult zebrafish. Transcriptomic analyses show that the regulated genes encompass most of the genes and pathways identified in previous studies (igfbp1, mTor, Notch, etc.), underlying the importance of those actors in beta cell regeneration. Our data reveal also that DNA replication is the most enriched signature attesting that duct cells undergo a potent proliferative response after the destruction of beta cells.

Besides these expected signatures, our transcriptomic data uncover the unanticipated upregulation of numerous genes implicated in DNA repair and cell cycle arrest. These signatures might indicate that highly proliferating ductal cells activate counteracting mechanisms. Among the genes regulated in those signatures, we focused on CaN and determined its role in beta cell regeneration. In an experimental setting in the young larvae revealing regeneration from the ducts [33], pharmacological inhibition of CaN increases the proliferation of duct cells induced by beta cell ablation, resulting in an acceleration of beta cell regeneration in the ducts. Consistently, we also used a genetic approach, and showed that transgene-mediated CaN overactivation abolishes the regenerative response. Importantly, the inhibition of regeneration is observed when CaN is overexpressed either ubiquitously or selectively in cftr-expressing ductal cells indicating that the role of CaN in beta cell regeneration is intrinsic to the ducts. Moreover, we showed that CaN acts on the same pool of ductal progenitors than Notch pathway and together control their proliferation and differentiation to beta cells. Altogether, these experiments confirm that the increased of beta cells induced by CaN originate from the ducts in the pancreatic tail. Nonetheless, while these experiments provide strong evidences, performing lineage tracing would further reinforce these data. We also suggest that CaN could improve regeneration in the principal islet, however as the molecular mechanisms of regeneration from the extra pancreatic ducts are not Notch dependant, it would be interesting to investigate the mechanisms under CaN inhibition.

Based on functional assays in larvae, we not only confirm the activation of the proliferation of ductal cells soon after beta cell ablation, but also that the rate of progenitor proliferation is carefully controlled by CaN in order to achieve proper and timely regeneration of beta cells. Our data are consistent with earlier studies reporting a role of CaN in proliferation dynamics during fin regeneration. In the regenerating fin, low CaN activity is found in the proximal region of the blastema characterized by a high rate of proliferation and regeneration and its activity increases distally where lower proliferation is observed [36], [38], [39]. It was suggested that CaN control blastemal cell progeny divisions [36]. In human, the importance of CaN in proliferation is also highlighted in organ transplanted patients. When patients are treated with Cyclosporin A (i.e. the CaN inhibitor we used in this study) as immunosuppressive drug they indeed present an increased risk of skin cancer, notably due to keratinocyte senescence inhibition [46].

It has been shown that Notch inhibitory treatments switch progenitors from proliferative self-renewing to premature differentiation, leading to progenitor depletion [34], [35]. Our study reveals that this phenomenon is further exacerbated by CaN inhibition. Importantly, during normal larval development in absence of Notch inhibitory treatment, CaN does not affect basal ductal proliferation nor beta cell differentiation. Hence, Notch signalling has to be repressed to detect the effect of CaN on the progenitors, suggesting that CaN acts downstream of Notch pathway. In differentiating keratinocytes, CaN cooperates with Notch signalling to regulate p21/cdkn1a (which is upregulated in the ducts at 3 dpt), cell cycle withdrawal and differentiation [47]. These studies show that CaN acts in association with Notch signalling on progenitors proliferation and on their differentiation.

Based on our data, we build the model depicted in Figure 7A. Our study suggests that CaN acts in competent progenitor cells and that this competence is determined by Notch signalling. When Notch is repressed to a mild level, the progenitor enter into the cell cycle and acquire a pro-endocrinogenic competence [35]. CaN acts on these progenitors to tone down an excessive proliferation and avoid the exhaustion of these progenitors. CaN is therefore a guardian of the progenitor population. CaN inhibition both increases progenitor proliferation and induces their depletion, suggesting a switch to a symmetric division resulting in two daughter cells entering in endocrine differentiation.

Model of CaN action on ductal progenitors to regenerate beta cells.

A) Under physiological conditions, the behavior of the ductal progenitors is determined by Notch signaling. Calcineurin is active in these progenitors and enable a proper control between proliferation and differentiation. When CaN in repressed, more ductal progenitors enter in the cell cycle (2dpt) and switch to a mode of proliferation leading to differentiation of the two daughter cells (4dpt), as more pro-endocrine cells are formed. The result is a exhaustion of the progenitors and a premature beta cell differentiation (10dpt).

More precisely, a previous study in mice uncover the existence of three division modes of the pancreatic progenitors during embryonic development, i.e. symmetric self-renewing resulting in two progenitors cells; asymmetric resulting in a progenitor and a endocrine cell and symmetric differentiative resulting in two differentiated cells [48]. The authors actually show that the type of division is defined by the timing of induction of endocrine program by NEUROG3 [48]. Interestingly, NEUROG3 seems to be the link between proliferation and differentiation, and its expression is regulated by Notch signalling [49]. In zebrafish, endocrine differentiation is not induced by Neurog3 but by Ascl1b and Neurod1 [50]. Concerning that subject, CaN inhibition accelerates the formation of neurod1+ cells. Hence CaN could possibly act via the determination of the type of division i.e. symmetric vs asymmetric. This model is supported by previous observation in others systems. In stem cells and neuronal and hematopoietic progenitors, premature differentiation results from a switch in the mode of cellular division, from symmetric amplifying division to asymmetric differentiating division [51], [52]. Notch determines the choice between both types of divisions [53], [54].

Calcineurin is known to be implicated in cellular senescence [46]. Usually thought as negative regulators of development and cellular growth, DNA repair [55] and cellular senescence [56] appear to be required in both developmental and regenerative processes. Our transcriptomic data suggest that these mechanisms are required for beta cell regeneration. Therefore, it would be interesting to determine the contribution of these cellular processes in the ductal progenitors and determine if CaN acts via cellular senescence in this case.

Overall, this study brings new insights on beta cell regeneration and highlights the ductal progenitor cell cycle as a cornerstone in the process. Some studies report an increase of proliferation of some ductal cell population in diabetic patients [10], [57], implying a regenerative response. However, these ductal cells cannot efficiently reform the beta cell mass, suggesting a dormant mechanism of regeneration. As such, the balance between proliferation and induction of endocrine differentiation could be a key to improve beta cell neogenesis. However, as CaN is also important for beta cell function, this approach would require to be transient to induce neogenesis. Therefore, it should be combined with methods to induce beta cell proliferation to ultimately reconstitute the beta cell mass. Overall, this study brings a better understanding on the regulation of the balance between ductal progenitors proliferation and endocrine differentiation. These results should provide new hints to help improve regenerative competences in mammals.

Material and Methods

Key resources table

Zebrafish husbandry and generation of the Tg(hsp70:eGFP-P2A-ppp3ccaCA)ulg068 and Tg(UAS:eGFP-P2A-ppp2ccaCA)ulg069 zebrafish lines

Tg BAC(nkx6.1:eGFP)ulg004 [58] ; Tg(ins:NTR-P2A-mCherry)ulg034 [59] ; Tg(cftr:gal4) and Tg(tp1:VenusPest) were used. Zebrafish were raised in standard conditions at 28°C. All experiments were carried out in compliance with the European Union and Belgian law and with the approval of the ULiège Ethical Committee for experiments with laboratory animals (approval number: 2075).

The hsp70:GFP-P2A-ppp3ccaCA transgene has been generated by cloning a PCR fragment containing the Gateway vector pCR8/GW/TOPO. Firstly, we amplified the full length of ppp3ccaCA with primers IM369/IM380 and amplified GFP-P2A with overlapping regions with IM371/O226. The overlapping PCR used the primers O226/IM380. Then to obtain a truncated ppp3ccaCA lacking the calmodulin biding and the autoinhibitory domain, resulting in a constantly active form of calcineurin, we amplified the last fragments with IM370/O226 and cloned into PCR8 vector. The promoter was assembled by LR recombination with pE5-hsp70 into pDestTol2p2A from the Tol2kit [60]. Tg (hsp70:GFP-P2A-ppp3ccaCA) fish have been generated using the Tol2 mediated transgenesis [61]. The Tg(UAS:GFP-P2A-ppp3ccacA) has been generated by ligation (KLD kit, NEB) of PCR fragments GFP-P2A-ppp3ccacA (IM459/IM460) with the UAS sequences (IM457/IM458) in plasmid from (Distel 2009) and then inserted into pDestTol2p2A from the Tol2kit. Final constructions has been injected with transposase into wild type (WT) AB embryos.

Beta-cell ablation

Adults fish for RNA-sequencing experiment were treated with freshly prepared metronidazole (MTZ) (Sigma M1547) at 10mM with 0.2% DMSO in fish water. Control treatments consisted of fish water containing 0.2% DMSO. Fish were treated for 18 hours in the dark. Nifurpirinol (NFP) (32439, Sigma-Aldrich) stock solution was dissolved at 2.5 mM in DMSO. b-cell ablation in Tg(nkx6.1:eGFP); Tg(ins:NTR-P2A-mCherry) larvae was induced by treatment with 2.5 µM NFP in E3. Control treatments consisted of E3 containing 0.2% DMSO. Larvae were treated for 18 hours in the dark.

Drug treatments

Cyclosporine A (Selleckchem, S2286), CHIR99021 and LY411575 (Sigma-Aldrich, SML0506) stock solution were dissolved at 10mM in DMSO. Larvae treatment were respectively performed at 0.1 µM and 5µM in E3. Control treatments consisted of E3 containing the same amount of DMSO than drug treatment. Larvae were treated for 18 hours in the dark.

5-ethynyl-2′-deoxyuridine (EdU) incorporation assay

Zebrafish larvae were incubated in 4 mM EdU dissolved in E3 water for 8 hours, they were then directly euthanised in tricaine and fixed in 4% PFA. EdU was detected according to the protocol of Click-iT™ EdU Cell Proliferation Kit for Imaging, Alexa Fluor™ 647 (ThermoFisher C10340) after whole mount immunodetection.

Heat Shock

Successive heat shocks of 30 minutes and 12 hours apart were performed at 39°C for larvae and 37°C for juveniles and adults zebrafish.

Whole mount immunodetection

Larvae were euthanized in tricaine and fixed in 4% PFA at 4 °C for IHC. The digestive tract of juveniles was dissected prior immunodetection and kept in methanol for at least 18 hours. After depigmentation with 3% H2O2/1% KOH, larvae were permeabilised 30 min in PBS/ Triton X-100 and incubated for two hours in blocking buffer (4% goat serum/1% BSA/PBS/0.1% Triton X-100). Primary and secondary antibodies were incubated at 4 °C overnight.

Primary antibodies: Living Colors Polyclonal anti-mCherry/dsRed (rabbit, 1:500, Clontech 632496), anti-GFP (chicken, 1:1000),

Secondary antibodies: Alexa Fluor-488, −568, −633 (goat, 1:750, Molecular Probes).

Flow cytometry and FACS

The whole pancreas from 3-4 fish of Tg(nkx6.1:eGFP); Tg(ins:NTR-P2A-mCherry) adult fish (6–10 months old, males and females) were dissected, collected and washed in HBSS without Ca2+/Mg2+. Live cell dissociation was performed in Tryple Select 1x solution (GIBCO) supplemented with 100 U/mL collagenase IV (Life Technologies 17104-019) and 40 µg/mL proteinase K (Invitrogen, 25530031) for 10 min at 28 °C.

The GFP+ cells were selected on FACS Aria III and sorted under purity mode and after exclusion of the doublets. The purity of the sorted cells was confirmed by epifluorescence microscopy (∼95 %). Cells (about 1000-5000/fish depending on the cell type) were immediately lysed with 0.5% Triton X-100 containing 2U/µl RNAse inhibitor and stored at –80 °C.

mRNA sequencing of FACSed cells and bioinformatic analyses

cDNAs were prepared from lysed cells according to SMART-Seq2.0 (Picelli et al., 2014) for low input RNA sequencing and libraries were prepared with Nextera® DNA Library kit (Illumina). Independent biological replicates of each cell type sequenced using Illumina NextSeq500 and obtained ∼20 million 100 bp paired-end reads. Reads were mapped and aligned to the zebrafish genome GRCz11 from Ensembl gene annotation version using STAR version 2.6.1 (Dobin et al., 2013). Gene expression levels were calculated with featureCounts (http://bioinf.wehi.edu.au/featureCounts/) and differential expression determined with DESeq2 (Love, Huber, & Anders, 2014). Expression values are given as normalized read counts. Poorly expressed genes with mean normalized expression counts <10 were excluded from the subsequent analyses. DESeq2 uses Wald test for significance with posterior adjustment of P values (Padj) using Benjamini and Hochberg multiple testing. The differentially expressed (DE) genes identified with a Padj cutoff of 0.05 were submitted for GO analysis using WebGestalt tool (Liao, Wang, Jaehnig, Shi, &Zhang, 2019).

Confocal microscopy and image analysis

Images were acquired using Leica SP5 confocal microscope. We used ImageJ to count the cells and Imaris to do the pictures.

Glycemia measurement

Glycemia measurement were performed as described in [59].

Data availability

The sequences that support the findings are of this study have been deposit in Geo with the accessions codes GSE212124, https://www.ncbi.nlm.nih.gov/geo/info/linking.html. The authors declare that all other data supporting the findings of this study are available within the paper and its supplementary information files.

Authors contribution

L.M. conducted experiments, acquired and analysed data, analysed transcriptomic data, wrote the manuscript. D.B. conducted RNA-seq experiment and revised the manuscript A.L. implemented bioinformatic programs and analysed transcriptomic data. C.R., C.G. and C.D. conducted experiments.

L.F. revised the manuscript. B.P. provided resources and advised on research studies. M.L.V. provided resources, advised on research studies and revised the manuscript I.M. supervised the whole work, acquired funding, designed the research studies, verified data and assisted in writing the manuscript.

Supporting information

Transcriptomic profiling of ductal cells during beta cell regeneration and validation in larvae

A) Design of the experimental pipeline for transcriptomic experiment

B) Expression of ppp3cca; ppp3ccb; nfatc3a; nfatc3b in acinar, alpha, beta, delta or ductal cells population from the zebrafish pancreas [40].

C) Calcineurin (ppp3ccb and ppp3cca) and NFATc3 (nfatc3a and nfatc3b) expression in ductal cells from zebrafish in non-ablated and ablated conditions.

Calcineurin inhibition with CsA increases the ductal regenerative response

A) Whole mount fluorescent immunohistochemistry (mCherry) of the pancreas of Tg(ins:NTR-P2A-mCherry) larvae at 4-7 and 14dpt. 3D projection (stack) of ablated larvae treated with DMSO or CsA representative samples. The principal islet (PI) and the pancreatic tail are shown. Scale 50 or 100µM.

B-C) Barplot representing the number of mCherry+ beta cells in the principal islet (PI) (B) and the number of secondary islets of mCherry+ beta cells in the pancreatic tail (C) of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 10dpt. Data are presented as mean values ± SD. T-test or Two-way ANOVA with Tukey’s multiple comparison test. **p-value<0.005

D) Barplot representing the number of sst1.1+ mCherry+ bi-hormonal cells in the principal islet (PI) of Tg(ins:NTR-P2A-mCherry); Tg(sst1.1:GFP) larvae at 6dpt. Data are presented as mean values ± SD. T-test.

E) Experimental design for EdU assay in larvae. After NFP treatment for 3 to 4dpf, larvae were exposed to EdU before fixation for analysis.

F) Whole mount fluorescent immunohistochemistry (GFP and EdU) of the pancreas of Tg(ins:NTR-P2A-mCherry); Tg(neurod1:GFP) larvae at 2-4-7 and 10 dpt. 3D projection (stack) of non-ablated and ablated larvae treated with DMSO or CsA representative samples. The principal islet (PI) and the pancreatic tail are shown. Scale 50µM.

G-H) Barplot representing the number of gcg+ alpha cells (G); the number of GFP+ sst1.1 delta cells (H); of Tg(ins:NTR-P2A-mCherry); Tg(sst1.1:GFP) larvae at 10dpt. Gcg was detected by IHC. The gray spheres represent non-ablated condition; the pink triangles represent the ablated condition; the black squares CsA condition and inverted green triangles ablated + CsA condition. Data are presented as mean values ± SD. Two-way ANOVA with Tukey’s multiple comparison test. **p-value<0.005

CaN repression potentializes the effect of Notch inhibition on beta cell formation

A) Experimental design for Notch inhibition test in non-ablated condition. Larvae were treated concomitantly with LY411575 (Notch inhibitor) and FK506 from 3 to 4dpf and were fixed and analyzed at 5dpf.

B) Barplot representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 5dpf. The black dots represent the control; gray FK506 treatment;blue LY411575; and purple combination of LY411575 and FK506. Data are presented as mean values ± SD. Two-way ANOVA with Tukey multiple comparison test, *p-value<0.05,**p-value<0.005, ****p-value<0.00005.

C) Experimental design for Notch inhibition test in non-ablated condition. Larvae were heat-shocked and then directly treated with LY411575 (Notch inhibitor) from 3 to 4dpf and were fixed and analyzed at 6dpf.

D) Barplot representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(hsp70:CaNCA); Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 5dpf. The black dots represent the control; gray heat-shock;blue LY411575; and purple combination of LY411575 and heat-shock. Data are presented as mean values ± SD. Two-way ANOVA with Tukey multiple comparison test, ****p-value<0.00005.

E) Barplot representing the mean number of mCherry+ beta cells in the pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 6dpf after Notch inhibition (LY411575), CaN inhibition (CsA) and NFATc inhibition (CHIR99021). Data are presented as mean values ± SD. Two-way ANOVA with Tukey multiple comparison test, *p-value<0.05,**p-value<0.005, ****p-value<0.00005.

CaN repression increases the proportion of duct proliferating cells

A) Z-plan pictures of Tg (nkx6.1:GFP); Tg (ins:NTR*-mCherry) larvae that were treated at 3dpf with both CsA 1µM and LY411575 5µM. At 4dpf, the larvae were exposed to EdU 4mM during 8 hours, before analysis at 6 dpf.

B-C) Barplot representing the number of GFP+ ductal cells which in pancreatic tail of Tg(ins:NTR-P2A-mCherry); Tg(nkx6.1:GFP) larvae at 4dpf for the Notch test. (B) Mild Notch inhibition with LY411575 5µM and (C) stronger Notch inhibition with LY411575 15µM. The black dots represent the control; gray CsA treatment;blue LY411575; and purple combination of LY411575 and CsA. Data are presented as mean values ± SD. T-test. Two-way ANOVA test with Tukey’s multiple comparisons test, *p-value<0.05; ****p-value<0.00005; ns = non-significant.

CaN regulation is important in juveniles/adults and necessary for correct glycemia recovery

A) Barplot representing the glycemia (mg/dL) of Tg(ins:NTR-P2A-mCherry); months old adult zebrafish at 14dpt. The pink triangles represent the ablated condition; the inverted green triangles ablated + CsA condition; the blue squares Tg(hsp70:CaNCA) after heat shocks. The grey line represents the mean glycemia of controls (non-ablated) fish. Data are presented as mean values ± SD.

B) Barplot representing the glycemia (mg/dL) of Tg(ins:NTR-P2A-mCherry); and Tg(UAS:CaNCA); Tg(cftr:gal4) in non-ablated adult zebrafish. Data are presented as mean values ± SD.